---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
```

```{r}
data <- read_csv("../data/tarca_et_al_2021/merged_normalised_datasets_for_modelling.csv")
data %>%
  filter(GEO_accession != "GSE150167") %>%
  count(Individual, GA) %>%
  count(Individual) %>%
  arrange(n)
```

```{r}
data %>%
  filter(GEO_accession == "GSE150167") %>%
  count(Log2RFU) %>% # range: 3.26 - 19.26
  arrange(desc(Log2RFU))

data %>%
  filter(GEO_accession != "GSE150167") %>%
  count(Log2RFU) %>% # range: 5.88 - 19.19
  arrange(desc(Log2RFU))
```

# Structure of raw data and quality control
```{r}
# How many gestational age measurements?
data %>%
  count(GA)

# Divide dataset into trimesters
data <- data %>% 
  mutate(Trimester = case_when(
    GA <= 13 ~ "First trimester",
    GA > 13 & GA <= 27 ~ "Second trimester",
    GA > 27 ~ "Third trimester"
  ))

data %>%
  count(Trimester)
```

```{r}
data %>%
  group_by(Trimester) %>%
  count(Group)

# Remove first trimester as it only contains controls
data <- data %>%
  filter(Trimester != "First trimester")

second_trimester <- data %>%
  filter(Trimester == "Second trimester")

third_trimester <- data %>%
  filter(Trimester == "Third trimester")
```

```{r}
# How many patients in the second trimester? 
second_trimester %>%
  count(Accession) %>%
  distinct(Accession) # 372

# What is the distribution of outcomes?
second_trimester %>%
  count(Group) # majority controls

# How many distinct targets were measured in the second trimester?
second_trimester %>%
  count(Target) %>%
  unique() %>% # 1,166 
  count(n) # (n varies between 105, 267, 372, 639, 906, 1173, 2241, 2508 and 9078)

# Identify targets with higher measurements and see how many measurements per patient
second_trimester %>%
  count(Target) #%>%
  #filter(n == 9078) # No Protein (??? remove)
  #filter(n == 2508) # HSP 70
  #filter(n == 2241) # Tenascin
  #filter(n == 1173) # GPNMB, IGFBP-2, Macrophage scavenger receptor
  #filter(n == 906) # JNK2
  #filter(n == 639) # ABL2
  #filter(n == 372) # ABL1
  #filter(n == 267) # ANR63
  #filter(n == 105) # BARK1
  
```

*The reason for the variation in n*
>Maternal plasma protein abundance was determined by using the SOMAscan platform v4.1, which is based on SOMAmer (slow off-rate modified aptamer) reagents. This platform allowed multiplexed quantification of 7288 analytes, corresponding to 6596 unique human protein targets in maternal plasma samples. Of these, 88.7% of proteins had one single assay, 10.4% had two, and less than 1% of proteins had 3–9 assays present on the platform ([Tarca et al., 2022](https://pubs.acs.org/doi/10.1021/acs.jproteome.2c00391)).

To do:
- remove "No Protein" target (9,078 rows in total)
- remove duplicates: targets with high n (line 59) have up to 9 samples taken in GSE206454 study (e.g. HSP 70), but all recorded at the same gestational age... useless

```{r}
second_trimester %>%
  filter(Target != "No Protein") %>%
  distinct(Accession, Target, .keep_all = TRUE) #%>% # drops from 505,809 rows to 408,888 rows -> not too bad!
  # How many subjects now?
  #count(Accession) %>% # 372 still, all good!
  # How many targets now?
  #count(Target) # 1165, all good here too!
```

Apply the changes and export the filtered dataset.
```{r}
second_trimester <- second_trimester %>%
  filter(Target != "No Protein") %>%
  distinct(Accession, Target, .keep_all = TRUE)

# How many measurements per target are there now?
second_trimester %>%
  count(Target) %>%
  unique() %>% 
  count(n) # Ranging from 105 to 267 to 372! 

second_trimester %>%
  filter(GEO_accession == "GSE206454") %>%
  count(GEO_accession, Individual) %>%
  count(n)

second_trimester %>%
  count(GA)
# 2178 - dupli uzorci (e.g. GA 13.1 i 21.1), 
# 3267 - tri uzorka (19.6, 23.6 i 26.6), 
# 4356 - 4 uzorka (14.6, 18.6, 22.6, 25.6)

#write_csv(second_trimester, "../data/tarca_et_al_2021/merged_second_trimester_normalised.csv")
```

Repeat the above steps for third trimester data frame now.
```{r}
# How many patients in the third trimester? 
third_trimester %>%
  count(Accession) %>%
  distinct(Accession) # 326

# What is the distribution of outcomes?
third_trimester %>%
  count(Group) # majority controls

# How many distinct targets were measured in the second trimester?
third_trimester %>%
  count(Target) %>%
  unique() %>% # 1,166, same as in the second trimester
  count(n) # (n varies between 105, 221, 326, 547, 768, 989, 1873, 2094 and 7514)

# Identify targets with higher measurements and see how many measurements per patient
third_trimester %>%
  count(Target) #%>%
  #filter(n == 7514) # No Protein (??? remove)
  #filter(n == 2094) # HSP 70
  #filter(n == 1873) # Tenascin
  #filter(n == 989) # GPNMB, IGFBP-2, Macrophage scavenger receptor
  #filter(n == 768) # JNK2
  #filter(n == 547) # ABL2
  #filter(n == 326) # ABL1
  #filter(n == 221) # ANR63
  #filter(n == 105) # BARK1
```

To do as above:
- remove "No Protein" target 
- remove duplicates
```{r}
third_trimester <- third_trimester %>% 
  filter(Target != "No Protein") %>%
  distinct(Accession, Target, .keep_all = TRUE)

# How many measurements per target are there now?
third_trimester %>%
  count(Target) %>%
  unique() %>% 
  count(n) # Ranging from 105 to 221 to 326! 

third_trimester %>%
  filter(GEO_accession == "GSE206454") %>%
  count(GEO_accession, Individual) %>%
  count(n)

third_trimester %>%
  count(GA)

#write_csv(third_trimester, "../data/tarca_et_al_2021/merged_third_trimester_normalised.csv")
```

Summary:
- removed first trimester samples as they contained only controls
- segregated the data into second and third trimester datasets
- removed duplicate target measurements per sample in both datasets

# Batch effects
```{r}
bioc_deps <- c("GO.db", "impute", "preprocessCore", "pvca","sva" )
cran_deps <- c("corrplot", "data.table", "ggplot2", "ggfortify","lazyeval", "pheatmap", "reshape2", "rlang", 
               "tibble", "dplyr", "tidyr", "wesanderson","WGCNA") 
               
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
#BiocManager::install(bioc_deps) 
#install.packages(cran_deps)

BiocManager::install(version = "3.18") 
BiocManager::install("proBatch")
```

# PCA visualisation
```{r}
second_trimester %>%
  select(Accession, Group, Target, Log2RFU) %>%
  pivot_wider(names_from = "Target", values_from = "Log2RFU") %>%
  column_to_rownames(var = "Accession")
```

```{r}
third_trimester %>%
  select(Accession, Group, Target, Log2RFU) %>%
  pivot_wider(names_from = "Target", values_from = "Log2RFU") %>%
  column_to_rownames(var = "Accession")
```

# Potential biomarkers
```{r}
potential_markers <- c("AGER", "APOB", "PTHLH", "YES1", "LAG3", "A2M", "LY86", "CDON", "SERPINA5", "CSNK2A1", "APP", "PLAUR", "PDE11A", "PCSK7", "EIF5", "SORCS2", "ITIH4", "IMPDH2", "JAK2")

data %>%
  select(Accession, EntrezGeneSymbol, GA, Log2RFU) %>%
  filter(Accession == "GSM6254314", EntrezGeneSymbol %in% potential_markers) %>%
  arrange(EntrezGeneSymbol)
```



















