---
title: "R Notebook"
output: html_notebook
---

```{r}
install.packages("minerva")
```

```{r}
library(tidyverse)
library(janitor)
library(ggpubr)
library(ggsci)
theme_set(
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12, colour = "black"),
    axis.title = element_text(size = 13),
    title = element_text(size = 14),
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5),
    panel.background = element_rect(fill = "white"),
    panel.grid = element_blank()
  )
)
```

```{r}
# Import the data
central_clusters <- read_csv("../output/tarca/central_clusters.csv")

hyp_res <- read_csv("../output/tarca/full_hypergraph_results.csv")
loeuf <- read_csv("../output/tarca/all_ranked_LOEUF_mapped_genes_control+sPTD.csv")

dunns <- read_csv("../output/tarca/statistics_filtered_DEGs_Dunns.csv")
anova <- read_csv("../output/tarca/statistics_filtered_DEGs_ANOVA.csv")
```

```{r}
# Separate the central clusters and DE results
central_clusters2 <- central_clusters %>%
  filter(trimester == "Second")
central_clusters3 <- central_clusters %>%
  filter(trimester == "Third")

dunns2 <- dunns %>%
  filter(trimester == "Second")
dunns3 <- dunns %>%
  filter(trimester == "Third")
anova2 <- anova %>%
  filter(trimester == "Second")
anova3 <- anova %>%
  filter(trimester == "Third")
```

```{r}
# Are all central cluster proteins DE?  
all(central_clusters2$target %in% anova2$target) # FALSE
all(central_clusters3$target %in% anova3$target) # FALSE
all(central_clusters2$target %in% dunns2$protein_name) # FALSE
all(central_clusters3$target %in% dunns3$protein_name) # FALSE

# Which proteins are DE?
central_clusters2$DE_anova <- central_clusters2$target %in% anova2$target
central_clusters2$DE_kw <- central_clusters2$target %in% dunns2$protein_name
central_clusters3$DE_anova <- central_clusters3$target %in% anova3$target
central_clusters3$DE_kw <- central_clusters3$target %in% dunns3$protein_name

# Are the DE_anova and DE_kw values the same for all proteins? Yes -> make a single DE column then
identical(central_clusters2$DE_anova, central_clusters2$DE_kw)
identical(central_clusters3$DE_anova, central_clusters3$DE_kw)

# Make a single DE column 
central_clusters2$DE <- central_clusters2$DE_kw
central_clusters2 <- central_clusters2 %>%
  dplyr::select(-c(DE_anova, DE_kw))

central_clusters3$DE <- central_clusters3$DE_kw
central_clusters3 <- central_clusters3 %>%
  mutate(DE_anova = NULL, DE_kw = NULL)

# Re-join the central clusters 
central_clusters <- full_join(central_clusters2, central_clusters3)
central_clusters # there should be 881 rows

central_clusters2 %>%
  count(DE)

central_clusters3 %>%
  count(DE)

central_clusters %>%
  select(target, DE) %>%
  distinct() %>%
  count(DE)
```

```{r}
# Are all central cluster proteins evolutionarily constrained?
all(central_clusters$target %in% loeuf$protein) # No.

# Which proteins are constrained?
central_clusters$constrained <- central_clusters$target %in% loeuf$protein

central_clusters %>%
  count(target) %>%
  arrange(desc(n)) # total is 645!

# How many proteins are constrained?
central_clusters %>%
  select(target, constrained) %>%
  distinct() %>%
  count(constrained)
```

```{r}
# Where are the central cluster proteins positioned in the hypergraph?
central_cluster_prots <- as.list(central_clusters$target)
all(central_clusters$target %in% hyp_res$protein) # TRUE

hyp_res %>%
  dplyr::select(trimester, group, protein, norm_rank) %>%
  filter(protein %in% central_cluster_prots) %>%
  distinct(protein, .keep_all = TRUE) %>% # should be 645
  count(norm_rank) # range from 0.06 - 1.0

hyp_res %>%
  dplyr::select(trimester, group, protein, norm_rank) %>%
  filter(protein %in% central_cluster_prots) %>%
  #distinct(protein, .keep_all = TRUE) %>% # should be 645
  ggplot() + aes(x = norm_rank, colour = group, fill = group) + geom_density(key_glyph = "path") +
  scale_colour_npg() + scale_fill_npg(alpha = 0.1) + facet_wrap(~trimester) + 
  ylab("Density") + xlab("Normalised rank")

ggsave("../output/tarca/visuals/cc_rank_density.tiff", height = 6, width = 8)

central_cluster_hyp <- hyp_res %>%
  dplyr::select(trimester, group, entrez_gene_symbol, protein, norm_rank) %>%
  filter(protein %in% central_cluster_prots) %>%
  distinct()

central_cluster_hyp %>%
  summary(norm_rank)

quantile(central_cluster_hyp$norm_rank, probs = 0.9)
quantile(central_cluster_hyp$norm_rank, probs = 0.1)
quantile(central_cluster_hyp$norm_rank, probs = 0.2)
quantile(central_cluster_hyp$norm_rank, probs = 0.3)
```

```{r}
central_cluster_hyp %>%
  filter(protein %in% c("Cyclophilin A", "IL-7 Ra", "FGF-12", "Livin B", "Arylsulfatase A", "GPC2", "DSC3"), group != "PPROM") %>%
  arrange(protein)

targets <- central_cluster_hyp %>%
  pivot_wider(id_cols = c(protein, trimester), 
              names_from = group,
              values_from = norm_rank) %>%
  select(-PPROM)

targets2 <- targets %>%
  filter(Control > 0.9197425 & sPTD < 0.1545064) # drop from the 90th quantile to the 10th quantile

central_cluster_hyp %>%
  filter(protein %in% targets2$protein, group != "PPROM") %>%
  ggplot() + aes(x = group, y = norm_rank, group = protein, colour = as.factor(protein)) + 
  geom_point(size = 3) + geom_line(size = 0.5, alpha  = 0.7) + facet_wrap(~trimester)

central_clusters %>%
  filter(target %in% targets2$protein) %>%
  count(constrained) # 11 DE, 25 non-DE, 28 constrained, 8 non-constrained

targets2$protein
```


```{r}
central_cluster_hyp2 <- left_join(central_cluster_hyp, central_clusters, by = c("trimester", "protein" = "target")) %>%
  drop_na()

cc_90th_quantile <- central_cluster_hyp2 %>%
  filter(norm_rank > 0.9197425) # 90th quantile

cc_90th_quantile %>%
  filter(DE == TRUE, constrained == TRUE)

cc_10th_quantile <- central_cluster_hyp2 %>%
  filter(norm_rank < 0.1545064) # 10th quantile

cc_10th_quantile %>%
  filter(DE == FALSE, constrained == FALSE)

#write_csv(cc_90th_quantile, "../output/tarca/central_cluster_90th_quantile.csv", col_names = TRUE)
#write_csv(cc_10th_quantile, "../output/tarca/central_cluster_10th_quantile.csv", col_names = TRUE)

top_50_10 <- cc_10th_quantile %>%
  filter(trimester == "Second") %>%
  arrange(norm_rank)

as.vector(top_50_10$protein)
```

# PCA loadings and DE proteins in the central cluster

```{r}
plotted <- c("JAK2", "SET", "GPC5", "UBE2N", "PCSK7", "NR1D1", "PA2G4", "PDPK1")
plotted %in% central_cluster_prots # [1]  TRUE FALSE FALSE  TRUE FALSE  TRUE FALSE FALSE

top_pos <- c('JAK2', 'IGFBP-6', 'NOTC2', 'PCSK7', 'C3a', 'METAP1', 'PKC-B-II', 'Integrin a1b1', 'FER', 'Clusterin', 'TSP4', 'IF4A3', 'GPC5', 'LG3BP', 'Glucagon', 'PTHrP', 'LYVE1', 'Coagulation Factor V', 'TIMP-1', 'LAG-3', 'M2-PK', 'CAMK2B', 'H6ST1', 'CAMK2D', 'Coagulation Factor VII')
top_pos %in% central_cluster_prots 

top_neg <- c('Thrombin', 'Soggy-1', 'a2-Macroglobulin', 'Cadherin-6', 'IL-17 RC', 'Prothrombin', 'SIRT2', 'OBCAM', 'YES', 'amyloid precursor protein', 'PIM1', 'MRCKB', 'C4b', 'P-Selectin', 'STX1a', 'Apo B', 'SPARCL1', 'Calcineurin B a', 'MMEL2', 'CSF-1', 'Flt-3', 'Protease nexin I', 'HIPK3', 'a1-Antichymotrypsin', 'GOT1')
top_neg %in% central_cluster_prots

central_clusters %>%
  filter(target == "HIPK3")

hyp_res %>%
  filter(protein == "STX1a")
```
*From the proteins plotted on Figure 3, JAK2, UBE2N and NR1D1 are in the central cluster.*
- JAK2 - term-specific in both trimesters, DE, constrained 
- NR1D1 - term-specific in both trimesters, DE, constrained
- UBE2N - sPTD-specific in both trimesters, DE, constrained
-> 3/8 (37.5%) are in the central cluster, and all three have known associations to premature birth.

*From the top positive loadings in second trimester, JAK2, Integrin a1b1, IF4A3, LG3BP, M2-PK, CAMK2B, CAMK2D, H6ST1 are in the central cluster.*
- JAK2 - term-specific in both trimesters, DE, constrained
- CAMK2B - sPTD-specific in third trimester, DE, constrained 
- CAMK2D - sPTD-specific in third trimester, DE, constrained 
- IF4A3 - term-specific in both trimesters, not DE, constrained
- LG3BP - sPTD-specific in third trimester, not DE, constrained
- H6ST1 - sPTD-specific in third trimester, not DE, constrained
- Integrin a1b1 - term-specific in both trimesters, not DE, not constrained
- M2-PK - term-specific in second trimester, not DE, not constrained
-> 8/25 (32%) are in the central cluster!

*From the top negative loadings in second trimester, Cadherin-6, IL-17 RC, YES, PIM1, C4b, STX1a, SPARCL1, Calcineurin B a, CSF-1, Flt-3, HIPK3 are in the central cluster.*
- PIM1 - term-specific in both trimesters, DE, constrained
- SPARCL1 - sPTD-specific in third trimester, DE, constrained
- HIPK3 - sPTD-specific in third trimester, DE, constrained
- Cadherin-6 - term-specific in third trimester, not DE, constrained
- IL-17 RC - term-specific in third trimester, not DE, constrained
- YES - sPTD-specific in third trimester, not DE, constrained
- STX1a - term-specific in both trimesters, not DE, constrained -> jumps from 90th to 10th percentile! Involved in Ca2+ regulation! 
- C4b - term-specific in both trimesters, not DE, not constrained
- Calcineurin B a - term-specific in both trimesters, not DE, constrained
- CSF-1 - term-specific in both trimesters, not DE, constrained
- Flt-3 - sPTD-specific in both trimesters, not DE, constrained
-> 11/25 (44%) are in the central cluster!

# Logistic modelling
```{r}
# Import the data
second_trimester <- read_csv("../data/tarca_et_al_2021/norm_log2t_batch_corrected_integrated_dataset_second_trimester.csv")
second_trimester

third_trimester <- read_csv("../data/tarca_et_al_2021/norm_log2t_batch_corrected_integrated_dataset_third_trimester.csv")
third_trimester

# Tidy the data
second_trimester$Trimester <- "Second"
third_trimester$Trimester <- "Third"

second_trimester <- second_trimester %>%
  clean_names()
third_trimester <- third_trimester %>%
  clean_names()

full_data <- full_join(second_trimester, third_trimester)
full_data <- full_data %>%
  clean_names()
```

```{r}
# Remove outliers
full_data_wide <- full_data %>%
  pivot_wider(id_cols = c(accession), names_from = target, values_from = log2rfu) %>%
  column_to_rownames(var = "accession")

set.seed(32)

# 1. Run PCA (if not already done)
pca <- prcomp(full_data_wide, center = TRUE, scale. = TRUE)

# 2. Decide on number of PCs (k)  
#    For example, capture 90% variance:
var_explained <- cumsum(pca$sdev^2) / sum(pca$sdev^2)
k <- which(var_explained >= 0.90)[1]

# 3. Extract the scores matrix (samples × k)
scores_k <- pca$x[, 1:k]

# 4a. Classical Mahalanobis in PC space
cov_k    <- cov(scores_k)
center_k <- colMeans(scores_k)
maha_k   <- mahalanobis(scores_k, center = center_k, cov = cov_k)

# 4b. Robust Mahalanobis via covMcd on PC scores
library(robustbase)
mcd_k    <- covMcd(scores_k, alpha = 0.75)
maha_robust_k <- mahalanobis(
  scores_k,
  center = mcd_k$center,
  cov    = mcd_k$cov
)

# 5. Define cutoff and flag outliers
#    e.g. 97.5th percentile of chi-square(df = k)
cutoff <- qchisq(0.975, df = k)
outliers_classic <- which(maha_k > cutoff)
outliers_robust  <- which(maha_robust_k > cutoff)

# 6. Remove or inspect outliers
filtered_samples <- setdiff(seq_len(nrow(full_data_wide)), outliers_robust)
filtered_data <- full_data_wide[filtered_samples, ]
```

```{r}
filtered_data2 <- filtered_data %>%
  rownames_to_column(var = "accession") %>%
  pivot_longer(cols = -c(accession), names_to = "target", values_to = "log2rfu")

filtered_data3 <- left_join(filtered_data2, full_data, by = c("accession", "target", "log2rfu"))
```

```{r}
# Restructure the data into wide format
model_data2 <- full_data %>% # change to filtered_data3
  filter(target %in% cc_90th_quantile$protein, group != "PPROM", trimester == "Second") # change cc_90th_quantile to cc_10th_quantile

model_data3 <- full_data %>%
  filter(target %in% cc_90th_quantile$protein, group != "PPROM", trimester == "Third")

model_data2w <- model_data2 %>%
  pivot_wider(#id_cols = c(accession, group), # use for logistic modelling
              id_cols = accession, # use for assessing multicollinearity and nonlinearity
              names_from = target,
              values_from = log2rfu) %>%
  column_to_rownames(var = "accession")

model_data3w <- model_data3 %>%
  pivot_wider(#id_cols = c(accession, group),
              id_cols = accession,
              names_from = target,
              values_from = log2rfu) %>%
  column_to_rownames(var = "accession")
```

*Assess multicollinearity*
Severe multicollinearity is present -> dimensionality reduction + regularization needed
- PCA, clustering or other orthogonalization techniques are required before biomarker modelling
- next steps: quantify collinearity using variance inflation factor (VIF) for individual proteins
```{r}
# Correlation matrix
library(corrplot)

cor_matrix2 <- cor(model_data2w, use = "pairwise.complete.obs")
print(cor_matrix2)

cor_matrix3 <- cor(model_data3w, use = "pairwise.complete.obs")
print(cor_matrix3)

png("../output/tarca/visuals/corrplot_cc10th_second_trimester.png", width = 800, height = 800)  
corrplot(cor_matrix2, method = "color", is.corr = TRUE, tl.pos = "n", order = "AOE")
dev.off()                                       

png("../output/tarca/visuals/corrplot_cc10th_third_trimester.png", width = 800, height = 800) 
corrplot(cor_matrix3, method = "color", is.corr = TRUE, tl.pos = "n", order = "AOE")
dev.off() 
```
```{r}
# Find pairs with |correlation| > 0.9
high_cor_pairs <- which(abs(cor_matrix2) > 0.75 & abs(cor_matrix2) < 1, arr.ind = TRUE)
cor_pairs_df <- data.frame(
  Protein_A = rownames(cor_matrix2)[high_cor_pairs[,1]],
  Protein_B = colnames(cor_matrix2)[high_cor_pairs[,2]],
  Correlation = cor_matrix2[high_cor_pairs]
)

model_data2w <- model_data2w[, !(colnames(model_data2w) %in% cor_pairs_df$Protein_A)]

high_cor_pairs3 <- which(abs(cor_matrix3) > 0.75 & abs(cor_matrix3) < 1, arr.ind = TRUE)
cor_pairs_df3 <- data.frame(
  Protein_A = rownames(cor_matrix3)[high_cor_pairs3[,1]],
  Protein_B = colnames(cor_matrix3)[high_cor_pairs3[,2]],
  Correlation = cor_matrix3[high_cor_pairs3]
)

model_data3w <- model_data3w[, !(colnames(model_data3w) %in% cor_pairs_df3$Protein_A)]
```

*Visualise nonlinear structure*
```{r}
library(Rtsne)
set.seed(42)
tsne_out <- Rtsne(as.matrix(model_data2w), dims = 2)
plot(tsne_out$Y, pch = 19, col = "steelblue", main = "t-SNE plot of samples")
```

```{r}
library(umap)
umap_out <- umap(model_data2w)
plot(umap_out$layout, pch = 19, col = "darkgreen", main = "UMAP plot of samples")
```

```{r}
filtered_data3 %>%
  filter(target %in% cc_10th_quantile$protein, group != "PPROM") %>%
  filter(target == "40S ribosomal protein SA") %>%
  ggplot() + aes(x = log2rfu, y = ga_delivery, colour = group) + geom_point()
```

```{r}
library(glmnet)

# Prepare data for glmnet
X <- model.matrix(~ . - 1, data = model_data2w[, -which(names(model_data2w) %in% c("group", "accession"))])
y <- model_data2w$group

# Check the reference group
model_data2w$group <- as.factor(model_data2w$group)
levels(model_data2w$group)

# Set or confirm the reference group if needed
model_data2w$group <- relevel(model_data2w$group, ref = "Control")  # "Control" is your control group label

# If binary outcome ("term"/"preterm"):
cv_fit <- cv.glmnet(X, y, family = "binomial", alpha = 0) # 'alpha = 0' for L2 penalty, 1 for L1 (Lasso), and 0.5 for elastic net (L1 + L2)

plot(cv_fit)
coef(cv_fit, s = cv_fit$lambda.1se)

cv_fit$lambda.min
cv_fit$lambda.1se 

better_lambda = cv_fit$lambda.1se 
```
*lambda.1se = lambda.min* -> what does this mean?
- cross-validation error is near minimum => the best (lambda min) and the simplest model (lambda 1se) are the same in performance
- indicates well-behaved structure of the data and the model

As the binomial deviance (a measure of cross-validated model fit; lower = better) plateaus at about X = 5, the additional number of predictors after that point does not meaningfully improve the model fit (as the curve flattens), therefore, instead of using the lambda.min for regularization, I will use the value of lambda found at X = 5, or 30 predictors, which is lambda = 0.0076303614.

```{r}
# Extract number of non-zero coefficients (including intercept) for each lambda
nonzero_counts <- cv_fit$glmnet.fit$df  # number of nonzero coeffs excluding intercept

# Lambda sequence tested
lambda_seq <- cv_fit$glmnet.fit$lambda

# View number of predictors (#nonzero) vs lambda
data.frame(lambda = lambda_seq, nonzero = nonzero_counts)
```


```{r}
# Extract coefficients at best lambda
coefs <- coef(cv_fit, s = better_lambda)

# Convert to a regular matrix and get row names
coefs_mat <- as.matrix(coefs)
predictor_names <- rownames(coefs_mat)

# Logical vector of non-zero coefficients (excluding intercept if you want)
nonzero <- coefs_mat[, 1] != 0

# Show only predictors with non-zero coefficients
nonzero_coefs <- coefs_mat[nonzero, , drop = FALSE]

# Display results
print(nonzero_coefs)
```
*What does this output represent?*
>The value of the coefficient represents the magnitude and direction of the effect of that predictor on the outcome on a specific scale (e.g., for logistic regression, it's on the log-odds scale). Each coefficient shows how log2rfu expression of a given biomarker affects the log-odds of being in the preterm group relative to control.

- positive coef -> higher log2rfu of that protein increases the odds of sPTD compared to the control
- negative coef -> higher log2rfu of that protein decreases the odds of sPTD compared to the control

An increase of one unit in log2rfu multiplies the odds by exp(coefficient).

```{r}
# Extract coefficients at lambda.min
coefs <- coef(cv_fit, s = better_lambda)

# Convert to matrix and extract names and values
coefs_mat <- as.matrix(coefs)
predictor_names <- rownames(coefs_mat)
coef_values <- coefs_mat[, 1]

# Filter non-zero coefficients
nonzero_idx <- coef_values != 0
nonzero_coefs <- coef_values[nonzero_idx]
nonzero_names <- predictor_names[nonzero_idx]
clean_names <- nonzero_names[nzchar(nonzero_names)]
print(clean_names)

# Calculate odds ratios
odds_ratios <- exp(nonzero_coefs)

# Calculate percentage increase/decrease in odds
# (odds ratio - 1) * 100 gives % change in odds
percent_change_odds <- (odds_ratios - 1) * 100

# Create a data frame for easy inspection
regularised_results4 <- data.frame( # change to (non)regularized if adding/removing regularization
  Predictor = nonzero_names,
  Coefficient = nonzero_coefs,
  OddsRatio = odds_ratios,
  PercentChangeOdds = percent_change_odds
)
regularised_results4$Predictor <- gsub("`", "", regularised_results4$Predictor)

# View results
#print(non_regularised_results)
#print(regularised_results) # L2
#print(regularised_results2) # L1
#print(regularised_results3) # elastic net
print(regularised_results4) # L2, 10th percentile

#write.csv(non_regularised_results, "../output/tarca/non_regularised_log_regression_modeling_central_cluster_90th_quantile_second_trimester.csv")
#write.csv(regularised_results, "../output/tarca/L2_regularised_log_regression_modeling_central_cluster_90th_quantile_second_trimester_minus_outliers.csv")
#write.csv(regularised_results2, "../output/tarca/L1_regularised_log_regression_modeling_central_cluster_90th_quantile_second_trimester.csv")
#write.csv(regularised_results3, "../output/tarca/elastic_net_regularised_log_regression_modeling_central_cluster_90th_quantile_second_trimester.csv")
write.csv(regularised_results4, "../output/tarca/L2_regularised_log_regression_modeling_central_cluster_10th_quantile_second_trimester_minus_outliers.csv")
```

```{r}
regularised_results4 %>%
  arrange(desc(PercentChangeOdds))
```


# Gene Ontology
```{r}
library(clusterProfiler)
library(annotables)
library(org.Hs.eg.db)
library(enrichplot)
library(pathview)
```

```{r}
# Sort out the central clusters for Functional Analysis
ctrl2 <- central_cluster_hyp %>%
  filter(trimester == "Second", group == "Control")
ctrl3 <- central_cluster_hyp %>%
  filter(trimester == "Third", group == "Control")
sptd2 <- central_cluster_hyp %>%
  filter(trimester == "Second", group == "sPTD")
sptd3 <- central_cluster_hyp %>%
  filter(trimester == "Third", group == "sPTD")

idx_sig <- grch38$symbol %in% sptd3$entrez_gene_symbol
ids_sig <- grch38[idx_sig, ]
non_duplicates2 <- which(duplicated(ids_sig$symbol) == FALSE)
ids_sig <- ids_sig[non_duplicates2, ]
sig_res_ids <- inner_join(sptd3, ids_sig, by=c("entrez_gene_symbol"="symbol"))
sig_res_ids <- sig_res_ids %>%
  distinct(entrez_gene_symbol)

# Create a background dataset for hypergeometric testing using all genes tested for significance in the results
all_genes <- as.character(hyp_res$entrez_gene_symbol)

# Significant results
sig_genes <- as.character(sig_res_ids$entrez_gene_symbol)
```

```{r}
# Biological functions
ego <- enrichGO(gene = sig_genes,
                #universe = all_genes, 
                keyType = "SYMBOL",
                OrgDb = org.Hs.eg.db,
                ont = "BP",
                pAdjustMethod = "BH",
                qvalueCutoff = 0.01,
                readable = TRUE)

#BP <- data.frame(ego) %>%
#  dplyr::filter(p.adjust < 0.01) %>%
#  arrange(desc(Count))
#BP$Trimester <- "Second"
#BP$Group <- "Control"

# BP2 <- data.frame(ego) %>%
#   dplyr::filter(p.adjust < 0.01) %>%
#   arrange(desc(Count))
# BP2$Trimester <- "Second"
# BP2$Group <- "sPTD"
# 
# BP_second_trimester <- full_join(BP, BP2)
# BP_second_trimester <- BP_second_trimester %>%
#   group_by(Group) %>%
#   arrange(desc(Count))
# 
# BP_second_trimester %>%
#   group_by(Description) %>%
#   wilcox.test(Group, FoldEnrichment, alternative = "two.sided", conf.int = TRUE, conf.level = 0.95)
# 
# write_csv(BP_second_trimester, "../output/tarca/biological_processes_central_clusters2.csv", col_names = TRUE)

# BP3 <- data.frame(ego) %>%
#   dplyr::filter(p.adjust < 0.01) %>%
#   arrange(desc(Count))
# BP3$Trimester <- "Third"
# BP3$Group <- "Control"

BP4 <- data.frame(ego) %>%
  dplyr::filter(p.adjust < 0.01) %>%
  arrange(desc(Count))
BP4$Trimester <- "Third"
BP4$Group <- "sPTD"

BP_third_trimester <- full_join(BP3, BP4)
BP_third_trimester <- BP_third_trimester %>%
  group_by(Group) %>%
  arrange(desc(Count))

#BP_third_trimester %>%
#  group_by(Description) %>%
#  wilcox.test(Group, FoldEnrichment, alternative = "two.sided", conf.int = TRUE, conf.level = 0.95)

write_csv(BP_third_trimester, "../output/tarca/biological_processes_central_clusters3.csv", col_names = TRUE)
```

```{r}
# Molecular functions
ego1 <- enrichGO(gene = sig_genes,
                #universe = all_genes, 
                keyType = "SYMBOL",
                OrgDb = org.Hs.eg.db,
                ont = "MF",
                pAdjustMethod = "BH",
                qvalueCutoff = 0.01,
                readable = TRUE)

#MF <- data.frame(ego1) %>%
#  dplyr::filter(p.adjust < 0.01) %>%
#  arrange(desc(Count))
#MF$Trimester <- "Second"
#MF$Group <- "Control"

# MF2 <- data.frame(ego1) %>%
#   dplyr::filter(p.adjust < 0.01) %>%
#   arrange(desc(Count))
# MF2$Trimester <- "Second"
# MF2$Group <- "sPTD"
# 
# MF_second_trimester <- full_join(MF, MF2)
# MF_second_trimester <- MF_second_trimester %>%
#   group_by(Group) %>%
#   arrange(desc(Count))
# 
# write_csv(MF_second_trimester, "../output/tarca/molecular_functions_central_clusters2.csv", col_names = TRUE)

# MF3 <- data.frame(ego1) %>%
#   dplyr::filter(p.adjust < 0.01) %>%
#   arrange(desc(Count))
# MF3$Trimester <- "Third"
# MF3$Group <- "Control"

MF4 <- data.frame(ego1) %>%
  dplyr::filter(p.adjust < 0.01) %>%
  arrange(desc(Count))
MF4$Trimester <- "Third"
MF4$Group <- "sPTD"

MF_third_trimester <- full_join(MF3, MF4)
MF_third_trimester <- MF_third_trimester %>%
  group_by(Group) %>%
  arrange(desc(Count))

# MF_third_trimester %>%
#   group_by(Description) %>%
#   wilcox.test(Group, FoldEnrichment, alternative = "two.sided", conf.int = TRUE, conf.level = 0.95)

write_csv(MF_third_trimester, "../output/tarca/molecular_functions_central_clusters3.csv", col_names = TRUE)
```

```{r}
# Cellular components
ego2 <- enrichGO(gene = sig_genes,
                #universe = all_genes, 
                keyType = "SYMBOL",
                OrgDb = org.Hs.eg.db,
                ont = "CC",
                pAdjustMethod = "BH",
                qvalueCutoff = 0.01,
                readable = TRUE)

#CC <- data.frame(ego2) %>%
#  dplyr::filter(p.adjust < 0.01) %>%
#  arrange(desc(Count))
#CC$Trimester <- "Second"
#CC$Group <- "Control"

# CC2 <- data.frame(ego2) %>%
#   dplyr::filter(p.adjust < 0.01) %>%
#   arrange(desc(Count))
# CC2$Trimester <- "Second"
# CC2$Group <- "sPTD"
# 
# CC_second_trimester <- full_join(CC, CC2)
# CC_second_trimester <- CC_second_trimester %>%
#   group_by(Group) %>%
#   arrange(desc(Count))
# 
# write_csv(CC_second_trimester, "../output/tarca/cell_compartments_central_clusters2.csv", col_names = TRUE)

# CC3 <- data.frame(ego2) %>%
#   dplyr::filter(p.adjust < 0.01) %>%
#   arrange(desc(Count))
# CC3$Trimester <- "Third"
# CC3$Group <- "Control"

CC4 <- data.frame(ego2) %>%
  dplyr::filter(p.adjust < 0.01) %>%
  arrange(desc(Count))
CC4$Trimester <- "Third"
CC4$Group <- "sPTD"

CC_third_trimester <- full_join(CC3, CC4)
CC_third_trimester <- CC_third_trimester %>%
  group_by(Group) %>%
  arrange(desc(Count))

write_csv(CC_third_trimester, "../output/tarca/cell_compartments_central_clusters3.csv", col_names = TRUE)
```

```{r}
# Sort out the central clusters for Functional Analysis
ctrl2 <- central_cluster_hyp %>%
  filter(trimester == "Second", group == "Control")
ctrl3 <- central_cluster_hyp %>%
  filter(trimester == "Third", group == "Control")
sptd2 <- central_cluster_hyp %>%
  filter(trimester == "Second", group == "sPTD")
sptd3 <- central_cluster_hyp %>%
  filter(trimester == "Third", group == "sPTD")
```

```{r}
comparelist <- list(ctrl2$entrez_gene_symbol, sptd2$entrez_gene_symbol)
names(comparelist) <- c("Control","sPTD") 
cclust <- compareCluster(geneCluster = comparelist, 
               fun = enrichGO,
               OrgDb= org.Hs.eg.db,
               keyType = "SYMBOL",
               ont= "BP")
```

```{r}
dotplot(cclust, showCategory=10)

cclust_df <- as.data.frame(cclust)

# Convert GeneRatio from "58/576" to numeric
cclust_df <- cclust_df %>%
  mutate(GeneRatioNum = sapply(GeneRatio, function(x) {
    num <- as.numeric(sub("/.*", "", x))
    den <- as.numeric(sub(".*/", "", x))
    num / den
  }))

# Keep top N terms per cluster (optional)
df_top <- cclust_df %>%
  group_by(Cluster) %>%
  slice_max(order_by = Count, n = 10) %>%
  ungroup()

# Plot
ggplot(df_top,
       aes(x = Cluster,
           y = fct_reorder(Description, Count),
           size = Count,
           color = p.adjust)) +
  geom_point() +
  scale_color_gradientn(
    colours = c("#f7ca64", "#46bac2", "#7e62a3"),
    trans = "log10",
    guide = guide_colorbar(reverse = TRUE)
  ) +
  #scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 11.5) +
  labs(
    x = "Cluster",
    y = NULL,
    size = "Count",
    color = "Adjusted p-value",
    title = "GO Biological Process Enrichment"
  )

```









