---
title: "R Notebook"
output: html_notebook
---

```{r}
# Install packages
#install.packages("tidyverse")
#BiocManager::install("dendextend")
#BiocManager::install("gplots")
#BiocManager::install("plyr")
#BiocManager::install("BioQC")
install.packages("duckplyr")
```

```{r}
# Load packages
library(tidyverse)
library(dendextend)
library(gplots)
library(plyr)
library(BioQC)
library(ggpubr)
library(ggsci)
library(duckplyr)
```

```{r}
# Set the working directory
setwd("../Desktop/Masters/RP-2/hypergraph")
getwd()
```
Filtered Ivan's data to gestational age = day 6 (lines 39-40).
```{r}
# Import normalized and filtered data
## Ivan's data
#exp_data <- read_csv("../data/trophectoderm_X.csv")
#obs_data <- read_csv("../data/trophectoderm_obs.csv")
#var_data <- read_csv("../data/trophectoderm_var.csv")

#obs_data <- obs_data %>%
#  filter(age == "6")

## Hu et al., 2025 data
exp_data <- read_csv("../data/hu_et_al_2025/hu_amnion_X.csv")
colnames(exp_data)[colnames(exp_data) == "...1"] <- "CellID"
obs_data <- read_csv("../data/hu_et_al_2025/hu_amnion_obs.csv")
colnames(obs_data)[colnames(obs_data) == "...1"] <- "CellID"
var_data <- read_csv("../data/hu_et_al_2025/hu_amnion_var.csv")
colnames(var_data)[colnames(var_data) == "...1"] <- "CellID"
```

```{r}
# Filter out genes with low variance
exp_data

# Calculate variances
variances <- sapply(exp_data, var)
print(variances)

# Check object class
class(variances)
typeof(variances)

# Calculate summary stats
min(variances, na.rm = TRUE)
max(variances, na.rm = TRUE)
summary(variances, na.rm = TRUE)

hist(variances)

# Variance cut-off = mean (0.037520)
variance_cutoff <- exp_data[, which(variances > 0.037520)]

# Return the CellID column
variance_cutoff$CellID <- exp_data$CellID
variance_cutoff <- variance_cutoff %>%
  select(CellID, everything())
```

```{r}
# Convert cell type column from character to factor
obs_data$cell_type <- as.factor(obs_data$cell_type)
```

*Hypergraph notes*

>Finally, if use has the value "pairwise.complete.obs" then the correlation or covariance between each pair of variables is computed using all complete pairs of observations on those variables. This can result in covariance or correlation matrices which are not positive semi-definite, as well as NA entries if there are no complete pairs for that pair of variables. For cov and var, "pairwise.complete.obs" only works with the "pearson" method (cor() R Documentation).
-> all genes x all genes => gene correlation matrix
-> using Pearson

```{r}
# Assign samples to rownames
#exp_data <- exp_data %>%
#  column_to_rownames("CellID")

exp_data <- variance_cutoff %>%
  column_to_rownames("CellID")

# Run hypergraph
runHypergraph <- function(sample, cell_subtype) {
  
  cell_ids <- sample$CellID
  samples_X <- exp_data[rownames(exp_data) %in% sample$CellID, ]
  samples_X <- na.omit(samples_X)
  
  # Remove genes with SD = 0
  sd_scores <- apply(samples_X, 2, sd)
  samples_X_sd <- samples_X[, which(sd_scores > 0)]
  
  # Run correlation
  gene_cor <- cor(samples_X_sd, use = "pairwise.complete.obs", method = "spearman")
  
  # Binerise correlation values using SD of correlation matrix
  bin <- abs(gene_cor)
  bin[which(bin > sd(gene_cor))] <- 1
  bin[which(bin != 1)] <- 0 # hypergraph incidence matrix
  incidence_matrix = as.matrix(bin)

  # Generate hypergraph adjacency matrix by matrix multiplication
  hyp <- bin %*% t(bin) 
  adjacency_matrix = as.matrix(hyp)
  
  # Ranking
  rowsum_res <- rowSums(hyp)
  #hist(rowsum_res) # non-parametric
  row_sums_df <- as.data.frame(rowsum_res)
  colnames(row_sums_df)[1] <- "rowsum"
  row_sums_df$rank <- rank(row_sums_df$rowsum, ties.method = "average")
  row_sums_df$norm_rank <- (row_sums_df$rank)/nrow(row_sums_df) # Divide by the number of rows
  
  # Add the leiden cluster label
  row_sums_df$cell_subtype <- cell_subtype
  
  # Sort ranks
  row_sums_df <- row_sums_df[order(row_sums_df$rank, decreasing = TRUE), ]
  
  # Change rownames to column
  row_sums_df$gene <- rownames(row_sums_df)

  return(row_sums_df)
  #return(adjacency_matrix)
}

a <- obs_data %>%
  group_by(cell_subtype) %>%
  nest() %>%
  mutate(hypergraph = map2(data, cell_subtype, runHypergraph))

a
a$data[[1]]
a$hypergraph[[1]]
```

*Number of genes each TE cluster has [6 clusters] -> Ivan's data*
leiden
<dbl>
data
<list>
hypergraph
<list>
1	<tibble>	<data.frame [17,890 × 5]>		
15	<tibble>	<data.frame [17,297 × 5]>		
9	<tibble>	<data.frame [16,198 × 5]>		
11	<tibble>	<data.frame [17,751 × 5]>		
0	<tibble>	<data.frame [15,959 × 5]>		
6	<tibble>	<data.frame [13,273 × 5]>	

```{r}
# Save the full hypergraph results
b <- a %>%
  mutate(hypergraph = map(hypergraph, ~ rownames_to_column(.x, var = "genes") %>% select(-leiden))) %>%
  unnest(data) %>%
  unnest(hypergraph) %>%
  select(leiden, gene, rowsum, norm_rank) %>%
  unique()
b

write_csv(b, "../output/full_hypergraph_results_spearman.csv", col_names = TRUE)
```

```{r}
library(glue)

visualiseHypergraph <- function(leiden) {
  
  adj_matrix_l0 <- a$hypergraph[[leiden]]
  euc_dist <- dist(adj_matrix_l0)
  hc <- hclust(euc_dist, method = "complete")
  
  png(glue("../output/hypergraph_analysis/plots/{leiden}_heatmap.png"), width = 1600, height = 1600, res = 300)
  heatmap(adj_matrix_l0, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc),
        symm = TRUE, scale = "none", col = heat.colors(256))
  dev.off()
  
}

b <- a %>%
  filter(leiden != 0)

plots <- b %>%
  dplyr::select(-data) %>%
  mutate(heatmaps = map(leiden, visualiseHypergraph))
```

```{r}
b %>%
  ggplot() + aes(x = rowsum) + geom_density(fill = "lightblue", alpha = 0.5) + theme_minimal() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) + 
  labs(title = "Density Plot of DEG Row Sums", x = "Row Sum", y = "Density")
```

```{r}
ggdensity(b, 
          x = "norm_rank", 
          title = "Hypernetwork Normalised Rank Distribution",
          ylab = "Density",
          xlab = "Normalised Rank") + 
  scale_colour_npg() + scale_fill_npg() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```

```{r}
# Save the filtered hypergraph results
c <- b %>%
  filter(norm_rank > 0.8)

write_csv(c, "../output/filtered_hypergraph_results_spearman.csv", col_names = TRUE)

# 5 clusters now! -> 0, 1, 9, 11, 15; 6 was dropped
c %>%
  select(leiden) %>%
  unique()
```

```{r}
# Save and export hypergraph incidence and adjacency matrices
incidence <- a %>%
  transmute(leiden, incidence = map(hypergraph, "incidence_matrix")) %>%
  unnest(incidence)

adjacency <- a %>%
  transmute(leiden, adjacency = map(hypergraph, "adjacency_matrix")) %>%
  unnest(adjacency)

write_csv(incidence, "../output/hypergraph_analysis/hypergraph_incidence_matrix_spearman.csv", col_names = TRUE)
write_csv(adjacency, "../output/hypergraph_analysis/hypergraph_adjacency_matrix_spearman.csv", col_names = TRUE)
```

```{r}
walk2(
  .x = a$leiden,
  .y = a$hypergraph,
  .f = function(cluster_id, cluster_data) {
    write.csv(cluster_data$incidence_matrix,
              paste0("../output/hypergraph_analysis/spearman_incidence_matrix_cluster_", cluster_id, ".csv"),
              row.names = TRUE)
  }
)

# run this
walk2(
  .x = a$leiden,
  .y = a$hypergraph,
  .f = function(cluster_id, cluster_data) {
    write.csv(cluster_data$adjacency_matrix,
              paste0("../output/spearman_hypergraph_analysis/adjacency_matrix_cluster_", cluster_id, ".csv"),
              row.names = TRUE)
  }
)
```

*Heatmap visualisation*
- leiden cluster 4

Following LOEUF mapping, the leiden cluster 4 is the only cluster in which two genes have been mapped to significant LOEUF scores with canonical transcripts. 

```{r}
a
```

```{r}
adj_matrix_l0 <- a$hypergraph[[1]]
euc_dist <- dist(adj_matrix_l0)
hc <- hclust(euc_dist, method = "complete")

png("../output/TE_leiden_1_heatmap.png", width = 1600, height = 1600, res = 300)
heatmap(adj_matrix_l0, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc),
        symm = TRUE, scale = "none", col = heat.colors(256))
dev.off()

adj_matrix_l0 <- a$hypergraph[[2]]
euc_dist <- dist(adj_matrix_l0)
hc <- hclust(euc_dist, method = "complete")

png("../output/TE_leiden_15_heatmap.png", width = 1600, height = 1600, res = 300)
heatmap(adj_matrix_l0, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc),
        symm = TRUE, scale = "none", col = heat.colors(256))
dev.off()

adj_matrix_l0 <- a$hypergraph[[3]]
euc_dist <- dist(adj_matrix_l0)
hc <- hclust(euc_dist, method = "complete")

png("../output/TE_leiden_9_heatmap.png", width = 1600, height = 1600, res = 300)
heatmap(adj_matrix_l0, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc),
        symm = TRUE, scale = "none", col = heat.colors(256))
dev.off()

adj_matrix_l0 <- a$hypergraph[[4]]
euc_dist <- dist(adj_matrix_l0)
hc <- hclust(euc_dist, method = "complete")

png("../output/TE_leiden_11_heatmap.png", width = 1600, height = 1600, res = 300)
heatmap(adj_matrix_l0, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc),
        symm = TRUE, scale = "none", col = heat.colors(256))
dev.off()

adj_matrix_l0 <- a$hypergraph[[5]]
euc_dist <- dist(adj_matrix_l0)
hc <- hclust(euc_dist, method = "complete")

png("../output/TE_leiden_0_heatmap.png", width = 1600, height = 1600, res = 300)
heatmap(adj_matrix_l0, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc),
        symm = TRUE, scale = "none", col = heat.colors(256))
dev.off()
```

```{r}
str(adj_matrix_l0)
```


```{r}
library(pheatmap)
pheatmap(adj_matrix_l0,
         clustering_distance_rows = euc_dist,
         clustering_distance_cols = euc_dist,
         clustering_method = "complete",
         show_rownames = TRUE,
         show_colnames = TRUE,
         color = colorRampPalette(c("white", "blue"))(256))
```


