---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(dplyr)
library(ggpubr)
library(ggsci)
```

```{r}
getwd()
```

```{r}
hyp_res <- read_csv("../output/full_hypergraph_results_spearman.csv")
hyp_res

hyp_filtered <- read_csv("../output/filtered_hypergraph_results_spearman.csv")
hyp_filtered

loeuf <- read_tsv("../data/loeuf/gnomad.v4.1.constraint_metrics.tsv")
loeuf
```

Column of interest from the constraint metrics is `lof.oe_ci.upper`, representing LOEUF: upper bound of 90% confidence interval for o/e ratio for high confidence pLoF variants (lower values indicate more constrained) ([gnomad_constraint_readme](https://storage.googleapis.com/gcp-public-data--gnomad/release/4.1/constraint/README.txt)).
- according to that: < 0.6 -> significance threshold for constraint, the lower the more constrained

```{r}
# Filter LOEUF scores 
loeuf_filtered <- loeuf %>%
  select(chromosome, gene, transcript, canonical, lof.oe_ci.upper, lof.oe_ci.upper_bin_decile) %>%
  filter(lof.oe_ci.upper < 0.6, canonical == TRUE) %>%
  distinct()

# Map LOEUF scores to unfiltered hypergraph results
loeuf_mapped_all <- inner_join(loeuf_filtered, hyp_res, by = "gene") %>%
  distinct()

write_csv(loeuf_mapped_all, "../output/TE_all_rank_LOEUF_mapped_genes.csv", col_names = TRUE)

# Map LOEUF scores to filtered hypergraph results
loeuf_mapped_filtered <- inner_join(loeuf_filtered, hyp_filtered, by = "gene") %>%
  distinct()

write_csv(loeuf_mapped_filtered, "../output/TE_filtered_rank_LOEUF_mapped_genes.csv", col_names = TRUE)
```

```{r}
loeuf_mapped_all %>%
  filter(leiden == 0) %>%
  ggplot() + aes(x = rowsum, y = norm_rank) + geom_point() + theme_minimal()

# All genes are constrained with canonical transcripts
loeuf_mapped_all %>%
  count("lof.oe_ci.upper") %>%
  arrange(desc(lof.oe_ci.upper))

# How many genes are constrained with canonical transcripts?
loeuf_mapped_all %>%
  count(gene) %>%
  distinct(gene) #3356

loeuf_mapped_filtered %>%
  count(gene) %>%
  distinct(gene) #2257
```

*Visualisation of Hypernetwork Rowsums*
- rowsums are calculated from the hypernetwork adjacency matrix -> indicate how strongly interconnected a gene is to other genes
- as it's based on an adjacency matrix and not a binary incidence matrix, this represent the total weight of connections a gene has
- peak in the low rowsums -> indicates a sparse connection between plenty of genes
- long tails -> indicate that hub genes are strongly interconnected
- a bimodal shape -> two modules -> indicates two strongly connected gene groups

```{r}
# Hypernetwork Rowsums
ggdensity(loeuf_mapped_all, 
          x = "norm_rank", 
          #color = "LOEUF", fill = "LOEUF", # Removed this row
          title = "Hypernetwork Normalised Rank Distribution",
          ylab = "Density",
          xlab = "Normalised Rank") + 
  scale_colour_npg() + scale_fill_npg() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

ggsave("../output/TE_norm_rank_density.tiff", height = 7, width = 7)
```

```{r}
# Genetic Constraint
loeuf_mapped_all %>%
  drop_na(chromosome) %>%
  filter(gene %in% c("ZFHX3")) %>%
  ggdotplot(., "cell_type", "norm_rank",
            add = "boxplot",
            color = "gene", fill = "gene",
            title = "Genetic constraint in first-trimester placenta",
            ylab = "Normalised rank",
            xlab = "Cell type") +
  scale_colour_npg() + scale_fill_npg() + theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) + facet_wrap(~chromosome)

#ggsave("../output/hypergraph_analysis/plots/constraint_dotplot_genes_w_boxplot.tiff", height = 7, width = 10)

loeuf_mapped_all %>%
  #filter(LOEUF == "Constrained") %>%
  drop_na(chromosome) %>%
  ggdotplot(., "cell_type", "norm_rank",
            add = "boxplot",
            color = "transcript", fill = "transcript",
            title = "Genetic constraint in first-trimester placenta",
            ylab = "Normalised rank",
            xlab = "Cell type") +
  scale_colour_npg() + scale_fill_npg() + theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) + facet_wrap(~chromosome)

#ggsave("../output/hypergraph_analysis/plots/constraint_dotplot_transcripts_w_boxplot.tiff", height = 7, width = 10)

# Leiden
loeuf_mapped %>%
  #filter(LOEUF == "Constrained") %>%
  drop_na(chromosome) %>%
  ggdotplot(., "cell_type", "norm_rank",
            add = "boxplot",
            color = "gene", fill = "gene",
            title = "Genetic constraint in first-trimester placenta",
            ylab = "Normalised rank",
            xlab = "Cell type") +
  scale_colour_npg() + scale_fill_npg() + theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) + facet_wrap(~leiden)

#ggsave("../output/hypergraph_analysis/plots/leiden_constraint_dotplot_genes_w_boxplot.tiff", height = 7, width = 10)
```

Following LOEUF mapping, the leiden cluster 4 is the only cluster in which two genes have been mapped to significant LOEUF scores with canonical transcripts. 

```{r}
# Violin plot of ranks
loeuf_mapped_all %>%
  group_by(tissue) %>%
  compare_means(norm_rank ~ cell_type, data = ., method = "wilcox.test")
my_comparisons <- list(c("EVT", "STB", "CTB", "HB", "FB", "END", "MIC", "TE", "EPI", "ICM", "HYPO"))

loeuf_mapped_all %>%
  #filter(LOEUF == "Constrained") %>%
  ggviolin(., "cell_type", "norm_rank", 
           color = "tissue", 
           add = "boxplot", 
           title = "Genetic constraint in blastocyst and first-trimester placenta",
           ylab = "Normalised rank",
           xlab = "Cell type") + 
  scale_colour_npg() + scale_fill_npg() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()) + 
  stat_compare_means(comparisons = my_comparisons) #, label.y = c(123, 135, 147)) + scale_y_continuous(limits = c(60, 155))

#ggsave("../output/hypergraph_analysis/plots/constraint_violin_blastocyst_placenta.tiff", height = 7, width = 8)
```

```{r}

```



