---
title: "R Notebook"
output: html_notebook
---

```{r}
install.packages("ARTool")
install.packages("emmeans")
install.packages("dummies")
devtools::install_github("js229/Vennerable")
install.packages("ggvenn")
```

```{r}
library(tidyverse)
library(janitor)
library(rstatix)
library(ARTool)
library(emmeans)
library(glmnet)
library(reprex)
library(ggpubr)
library(ggsci)
library(scales)
library(correlation)
library(Vennerable)
library(ggvenn)
theme_set(
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12, colour = "black"),
    axis.title = element_text(size = 13),
    title = element_text(size = 14),
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5),
    panel.background = element_rect(fill = "white"),
    panel.grid = element_blank()
  )
)
```

```{r}
kmeans_clusters <- read_tsv("../data/tarca_et_al_2021/string/string_kmeans_clusters.tsv")
kmeans_clusters

cluster_desc <- read_tsv("../data/tarca_et_al_2021/string/string_kmeans_cluster_descriptions.tsv")
cluster_desc

second_trimester <- read_csv("../data/tarca_et_al_2021/norm_log2t_batch_corrected_integrated_dataset_second_trimester.csv")
second_trimester

third_trimester <- read_csv("../data/tarca_et_al_2021/norm_log2t_batch_corrected_integrated_dataset_third_trimester.csv")
third_trimester
```

```{r}
# Tidy the data
kmeans_clusters <- kmeans_clusters %>%
  clean_names()

second_trimester$Trimester <- "Second"
third_trimester$Trimester <- "Third"

second_trimester <- second_trimester %>%
  clean_names()
third_trimester <- third_trimester %>%
  clean_names()

full_data <- full_join(second_trimester, third_trimester)
full_data <- full_data %>%
  clean_names()
```

```{r}
third_trimester %>%
  select(accession, group) %>%
  unique() %>%
  count(group)

second_trimester %>%
  filter(group == 'Control') %>%
  arrange(desc(log2rfu))
```

```{r}
# How many proteins and samples in each data subset?
full_data %>%
  count(target) %>% 
  count(n) 
# 1165 targets in total, each measured in 698 patient samples (566 controls, 124 sPTDs, 8 pPROMs)
full_data %>%
  filter(group == "PPROM") %>%
  count(accession) # 566 controls, 124 sPTDs, 8 pPROMs

second_trimester %>%
  count(target) # 1165 proteins in 372 samples

third_trimester %>%
  count(target) # 1165 proteins in 326 samples

# Quality control
targets_ctr2 <- unique(second_trimester$target[second_trimester$group == 'Control'])
targets_spd2 <- unique(second_trimester$target[second_trimester$group == 'sPTD'])

# Find common values
common_values <- intersect(targets_ctr2, targets_spd2)

if (length(common_values) > 0) {
  cat("The two groups share", length(common_values), "common values.")
} else {
  cat("The two groups do not share any common values.\n")
}
```

```{r}
# Identify matching columns
kmeans_clusters %>%
  count(protein_name) # 1112 proteins

# Create lists to check for overlap -> use in stats testing based on clusters
tarca_list <- as.list(full_data$target)
cluster_list <- as.list(kmeans_clusters$protein_name)

count <- tarca_list %in% cluster_list
table(count)
```

```{r}
# Join the datasets
cluster_stats <- right_join(kmeans_clusters, full_data, by = c("protein_name" = "target"))
cluster_stats %>%
  count(protein_name) # 1,165 proteins

# Select only relevant columns
cluster_stats <- cluster_stats %>%
  select(cluster_number, protein_name, log2rfu, group, trimester)
```

# Normality testing
Log2RFU data follows a 'leptokurtic distribution' (more sharply peaked than normal dist)
-> may indicate low variance or strong central tendency
-> slightly left-skewed + bimodal: could reflect biological subgroups or measurement noise - investigate
-> distributions are similar across trimesters and outcome groups
```{r}
# Quantitative normality testing: Kolmogorov-Smirnov (for datasets with > 50 samples)
full_data %>%
  group_by(group) %>%
  summarise(
    n = n(),
    mean = mean(log2rfu),
    sd = sd(log2rfu),
    ks = list(ks.test(log2rfu, "pnorm", mean(log2rfu), sd(log2rfu)))
  ) %>%
  mutate(
    D = sapply(ks, function(x) x$statistic),
    p_value = sapply(ks, function(x) x$p.value)
  )
# p < 0.05: reject the null hypothesis -> not normally distributed
```

```{r}
# Install required packages
install.packages(c("fitdistrplus", "sn", "mixsmsn"))

library(fitdistrplus)
library(sn)        # for skewed distributions
library(mixsmsn)   # for mixture skewed-normal/t

# Example: suppose your data is in a vector called "x"
x <- full_data$log2rfu
# Fit skewed t-distribution
fit_st <- selm(x ~ 1, family = "ST")  # from 'sn' package

# Fit skewed Laplace
# (no direct base function; use VGAM or custom MLE)

# Fit mixture of skewed normals/t
fit_mix <- smsn.mix(x, nu = 4, g = 2, family = "Skew.t")  # 2-component skewed-t mixture

# Compare AIC values
AIC(fit_st)   # skewed t
fit_mix$aic   # mixture skewed-t
```

```{r}
pdf_values <- dsgt(x, mu = 1, sigma = 1, lambda = -0.3, p = 3, q = 100, mean.cent = FALSE)

# Create a data frame for plotting
plot_data <- data.frame(x = x, density = pdf_values)

# Create the plot
ggplot(plot_data, aes(x = x, y = density)) +
  geom_line(color = "blue", linewidth = 1) +
  labs(title = "Generalized Skewed t-Distribution PDF",
       x = "x",
       y = "Density") +
  theme_minimal() + 
  ylim(0, 1) + xlim(-5, 5)
```

```{r}
# Density curve
dc <- cluster_stats %>%
  ggplot() + aes(x = log2rfu, colour = group) + 
  geom_density(key_glyph = "path") + facet_wrap(~trimester) + 
  scale_colour_npg() + scale_fill_npg() + 
  labs(x = "Sample signal (log2RFU)", y = "Density")

# QQ plot 
qqp <- cluster_stats %>%
  ggplot() + aes(sample = log2rfu, colour = group) + 
  geom_qq() + geom_qq_line() + facet_wrap(~trimester) +
  scale_colour_npg() + scale_fill_npg() + 
  ylab("Sample quantiles") + xlab("Theoretical quantiles")

combined_plot <- ggarrange(dc, qqp, common.legend = TRUE, legend = "top")
annotate_figure(combined_plot, top = text_grob("Sample Signal Distribution Across Trimesters", size = 14))

ggsave("../output/tarca/visuals/density-and-qq.tiff", height = 4, width = 9)
```
```{r}
full_data %>%
  filter(target == "P-Selectin") %>% # P-Selectin
  ggplot() + aes(x = ga_delivery, y = log2rfu, colour = trimester) +
  geom_point() + facet_wrap(~group) +
  scale_colour_npg() + scale_fill_npg()
```


# Equal variance testing
Null hypothesis in Levene's test is that all groups have equal variances (homoscedasticity). 
```{r}
# Global variance test
full_data %>%
  levene_test(log2rfu~group*trimester, center = "median") # Brown-Forsythe test

# Per-trimester variance test
second_trimester %>%
  levene_test(log2rfu ~ group, center = "median")
third_trimester %>%
  levene_test(log2rfu ~ group, center = "median")
```
p < 0.05 in all cases -> reject the null hypothesis; the assumption of equal variances is violated, and therefore, the data in both trimester subsets is heteroscedastic. 

# Statistical evaluation based on clusters
##Kruskal-Wallis test
-> is there a significant difference in log2RFU between different pregnancy outcomes ('group') across the clusters in second and third trimester?
```{r}
doKruskalTest <- function(x) {
  x %>%
    kruskal_test(log2rfu~group) %>%
  print(kruskalTest)
}

a <- cluster_stats %>%
  group_by(cluster_number, trimester) %>%
  nest()

a$data[[1]]

a <- cluster_stats %>%
  group_by(cluster_number, trimester) %>%
  nest() %>%
  mutate(Kruskal = map(data, doKruskalTest))

a$Kruskal[[1]]
a$data[[1]]

cluster_res_kruskal <- a %>%
  unnest(cols = c("cluster_number", "trimester", "Kruskal")) %>%
  select(-data)

# Adjust p values
cluster_res_kruskal$p_adj_bh <- p.adjust(cluster_res_kruskal$p, method = "BH") # Benjamini–Hochberg correction
cluster_res_kruskal$p_adj_by <- p.adjust(cluster_res_kruskal$p, method = "BY") # Benjamini–Yekutieli correction

cluster_res_kruskal %>%
  filter(p < 0.05) # 17 significant overall 
cluster_res_kruskal %>%
  filter(p_adj_bh < 0.05) # 16 significant with BH correction
cluster_res_kruskal %>%
  filter(p_adj_by < 0.05) # 14 significant with BY correction

# Reorder the columns
cluster_res_kruskal <- cluster_res_kruskal %>%
  select(-method, everything(), method) %>%
  drop_na(cluster_number)

write_csv(cluster_res_kruskal, "../output/tarca/statistics_kruskal_wallis_results_clusters.csv")

# Next steps
dunns <- a %>%
  unnest(cols = c("cluster_number", "trimester", "data", "Kruskal"))
dunns$p_adj_by <- p.adjust(dunns$p, method = "BY") # Benjamini–Yekutieli correction
```
The answer is **yes.**

### Dunn's test
Which pregnancy groups differ across clusters in second and third trimester?
```{r}
doDunnsTest <- function(x) {
    x %>%
    dunn_test(log2rfu~group) %>%
  print(dunnsTest)
  }

b <- dunns %>%
  group_by(cluster_number, trimester) %>%
  dplyr::filter(p_adj_by < 0.05) %>%
  nest()

b$data[[1]]

b <- dunns %>%
  group_by(cluster_number, trimester) %>%
  dplyr::filter(p_adj_by < 0.05) %>%
  nest() %>%
  mutate(Dunn = map(data, doDunnsTest))

b$Dunn[[1]]

cluster_res_dunn <- b %>%
  unnest(cols = c("cluster_number", "trimester", "Dunn")) %>%
  select(-data) %>%
  drop_na(cluster_number) %>%
  arrange(p.adj)

write_csv(cluster_res_dunn, "../output/tarca/statistics_dunns_results_clusters.csv")
```

STRING cluster descriptions:
1 - Transmembrane receptor protein tyrosine kinase signalling pathway (411 genes)
2 - Positive regulation of immune system process (387 genes)
*6 - Neurexins and neuroligins (7 genes = NRXN3, NRXN1, SPARCL1, LRRTM1, NXPH1, NLGN4X, LRRTM3)*
- SPARCL1 much more expressed in sPTD in PCA clustering in second trimester - it is also the only protein out of these 7 in the PCA

>Neurexins and neuroligins are synaptic cell-adhesion molecules that connect presynaptic and postsynaptic neurons at synapses, mediate signalling across the synapse, and shape the properties of neural networks by specifying synaptic functions. In humans, alterations in genes encoding neurexins or neuroligins have recently been implicated in autism and other cognitive diseases, linking synaptic cell adhesion to cognition and its disorders ([Sudhof, 2008](https://www.nature.com/articles/nature07456)).

>Neurexins (NXs) and neuroligins (NLs) are transsynaptically interacting cell adhesion proteins that play a key role in the formation, maturation, activity-dependent validation, and maintenance of synapses. As complex alternative splicing processes in nerve cells generate a large number of NX and NLs variants, it has been proposed that a combinatorial interaction code generated by these variants may determine synapse identity and network connectivity during brain development. The functional importance of NXs and NLs is exemplified by the fact that mutations in NX and NL genes are associated with several neuropsychiatric disorders, most notably with autism ([Krueger et al., 2012](https://doi.org/10.1016/j.conb.2012.02.012)).

*8 - TNFs bind their physiological receptors (3 genes = EDA, EDA2R, EDAR)*


*10 - Protein K69-linked ufmylation (2 genes = UFC1, UFM1)*
-> entirely novel: there is nothing in relation to preterm birth + Protein K69-linked ufmylation in the literature!
> While physiological triggers and consequences of UFMylation are not entirely clear, its biological importance is epitomized by mutations in the UFMylation pathway in human pathophysiology including musculoskeletal and neurodevelopmental diseases ([Millrine et al., 2023](https://doi.org/10.1111/febs.16730)).

-> related to retinopathy of prematurity
> Cilia are evolutionarily conserved organelles characterized by a microtubule-based structure, known as the axoneme. The axoneme is surrounded by the ciliary membrane and originates from the basal body, a structure derived from the mother centriole. Cilia play a critical role in embryonic development, including organ morphogenesis and the establishment of the left-right asymmetry.
...
Mechanistic studies have revealed that ciliary abnormalities are often caused by abnormal expression or mislocalization of ciliary proteins, including those that constitute the structural components of cilia and those involved in maintaining ciliary homeostasis, such as ciliary assembly, disassembly, and length control.[4] These alterations in ciliary proteins are largely regulated by post-translational modifications (PTMs), which significantly enhance the complexity of the proteome and are crucial for maintaining tissue homeostasis.[5] Examples of PTMs involved in ciliary regulation include acetylation, glycylation, palmitoylation, and the recently identified ubiquitin-fold modifier 1 (UFM1) modification (UFMylation). These modifications play crucial roles in maintaining ciliary homeostasis and contribute to the pathogenesis of ciliopathies by modifying ciliary proteins.
...
Accumulating evidence has shown that dysfunction in these PTMs can lead to perturbations in tissue homeostasis, contributing to human diseases.[9] For example, [...] the interplay between the phosphorylation and ubiquitination of histone deacetylase 6 (HDAC6), a key regulator of ciliary disassembly, has been linked to retinopathy of prematurity through its role in maintaining the integrity of the photoreceptor cilium ([Ran and Zhou, 2025](https://doi.org/10.1002/advs.202416562)).

## Aligned-Rank Transform (ART) ANOVA
Calculate the proportion of tied values before conducting ART ANOVA as it can be sensitive to high ties (above 10-15%).
```{r}
n_ties <- sum(duplicated(full_data$log2rfu))
prop_ties <- n_ties / length(full_data$log2rfu) # No duplicated log2rfu values => no heavy ties
```

```{r}
# Ensure relevant character columns are stored as factors
cluster_stats$group <- factor(cluster_stats$group)
cluster_stats$trimester <- factor(cluster_stats$trimester)
cluster_stats$cluster_number <- factor(cluster_stats$cluster_number)
```

```{r}
# Fit the aligned-rank transform model
art_model <- art(log2rfu ~ group * trimester * cluster_number, data = cluster_stats)
summary(art_model)

# Run ANOVA on the model
anova_res <- anova(art_model)
print(anova_res)
```
1 factor(group)                                           2 174440 11.5417889 9.7229e-06 ***
-> strong evidence to reject the null hypothesis of no difference between groups
-> indicates median log2rfu differs between groups

2 factor(trimester)                                       1 174440  0.0017346    0.96678    
-> no evidence to reject the null hypothesis of no difference between trimesters

3 factor(cluster_number)                                  9 174440 64.3353861 < 2.22e-16 ***
-> strong evidence to reject the null hypothesis of no difference between clusters
-> indicates median log2rfu differs between clusters

4 factor(group):factor(trimester)                         2 174440  1.5746356    0.20709    
-> interaction of the two factors is insignificant, suggesting no evidence that group differences depend on trimeste
-> the effect of log2rfu does not changes depending on the trimester or sample-taking

5 factor(group):factor(cluster_number)                   18 174440 11.5696782 < 2.22e-16 ***
-> significant interaction of the two, suggesting strong evidence that group differences depend on clusters

6 factor(trimester):factor(cluster_number)                9 174440  0.6395194    0.76410    
-> interaction of the two factors is insignificant, suggesting no evidence that trimester differences depend on clusters

7 factor(group):factor(trimester):factor(cluster_number) 18 174440 11.7349438 < 2.22e-16 ***
-> interaction of the three is significant, suggesting evidence that the interaction between group and trimester changes across clusters

```{r}
posthoc <- art.con(art_model, "group", adjust = "bonferroni") # both bonferroni and by are available
summary(posthoc)

posthoc <- art.con(art_model, "cluster_number", adjust = "bonferroni")
summary(posthoc)

posthoc <- art.con(art_model, "group:cluster_number", adjust = "bonferroni")
summary(posthoc)

posthoc <- art.con(art_model, "group:trimester:cluster_number", adjust = "bonferroni")
summary(posthoc)

posthoc_df <- summary(posthoc)
posthoc_df <- as.data.frame(posthoc_df)
```

```{r}
# Build per-term linear model
lm_condition <- artlm(art_model, term = "group", response = "art")

# Summary for inference on the term
summary(lm_condition)
```

## Generalised Linear Modelling
```{r}
model <- glm(group ~ log2rfu + target, family = binomial(link = "logit"), data = full_data)
summary(model)
```

# Statistical evaluation based on proteins
## Kruskal-Wallis test
-> is there a significant difference in log2RFU between different pregnancy outcomes ('group') across proteins?
```{r}
doKruskalTest <- function(x) {
  x %>%
    kruskal_test(log2rfu~group) %>%
  print(kruskalTest)
}

c <- cluster_stats %>%
  group_by(protein_name, trimester) %>%
  nest()

c$data[[1]]

c <- cluster_stats %>%
  group_by(protein_name, trimester) %>%
  nest() %>%
  mutate(Kruskal = map(data, doKruskalTest))

c$Kruskal[[1]]
c$data[[1]]
c

# Explore the results
kruskal_res_proteins <- c %>%
  unnest("Kruskal") %>%
  select(-data)

kruskal_res_proteins$p_adj_bh <- p.adjust(kruskal_res_proteins$p, method = "BH") # Benjamini–Hochberg correction
kruskal_res_proteins$p_adj_by <- p.adjust(kruskal_res_proteins$p, method = "BY") # Benjamini–Yekutieli correction

kruskal_res_proteins %>%
  filter(trimester == "Third", p < 0.05) %>%
  count(protein_name) # 1086 proteins in second trimester + 1101 proteins in third trimester (out of 1165)

kruskal_res_proteins %>%
  filter(trimester == "Third", p_adj_bh < 0.05) %>%
  count(protein_name) # 1085 proteins in second trimester + 1096 proteins in third trimester (out of 1165)

kruskal_res_proteins %>%
  filter(trimester == "Third", p_adj_by < 0.05) %>%
  count(protein_name) # 1039 proteins in second trimester + 1048 proteins in third trimester (out of 1165)

# Reorder the columns
kruskal_res_proteins <- kruskal_res_proteins %>%
  select(-method, everything(), method) %>%
  arrange(p_adj_by)

kruskal_res_proteins %>%
  filter(trimester == "Third", p_adj_by < 0.05) %>%
  select(protein_name) %>%
  count(protein_name) %>%
  distinct(protein_name)

#write_csv(kruskal_res_proteins, "../output/tarca/statistics_kruskal_wallis_results_proteins.csv")

# Next steps
dunns2 <- c %>%
  unnest(cols = c("protein_name", "trimester", "data", "Kruskal"))
dunns2$p_adj_by <- p.adjust(dunns2$p, method = "BY") # Benjamini–Yekutieli correction
```
**yes!**

*Post-hoc: Dunn's test*
-> which pregnancy outcomes (groups) have statistically significant difference in log2RFU protein expression?
```{r}
doDunnsTest <- function(x) {
    x %>%
    dunn_test(log2rfu~group) %>%
  print(dunnsTest)
  }

d <- dunns2 %>%
  group_by(protein_name, trimester) %>%
  dplyr::filter(p_adj_by < 0.05) %>%
  nest()

d$data[[1]]

d <- dunns2 %>%
  group_by(protein_name, trimester) %>%
  dplyr::filter(p_adj_by < 0.05) %>%
  nest() %>%
  mutate(Dunn = map(data, doDunnsTest))

d$Dunn[[1]]

dunn_res_proteins <- d %>%
  unnest(cols = c("protein_name", "trimester", "Dunn")) %>%
  select(-data) %>%
  arrange(p.adj)

dunn_res_proteins %>%
  filter(trimester == "Third", p.adj < 0.05) %>% 
  arrange(p.adj) %>%
  count(protein_name) %>%
  unique() # 1018 proteins in second trimester + 1036 proteins in third semester
?dunn_test
#write_csv(dunn_res_proteins, "../output/tarca/statistics_dunns_results_proteins.csv")
```

## Aligned-Rank Transform ANOVA
```{r}
# Ensure relevant character columns are stored as factors
full_data$group <- factor(full_data$group)
full_data$trimester <- factor(full_data$trimester)
full_data$target <- factor(full_data$target)

second_trimester$group <- factor(second_trimester$group)
second_trimester$trimester <- factor(second_trimester$trimester)
second_trimester$target <- factor(second_trimester$target)

third_trimester$group <- factor(third_trimester$group)
third_trimester$trimester <- factor(third_trimester$trimester)
third_trimester$target <- factor(third_trimester$target)
```

```{r}
# Define ART ANOVA function
doARTAnova <- function(x) {
  if(length(unique(x$group)) < 2) {
    return(NA) # not enough groups to compare
  } else {
    art_model <- art(log2rfu ~ group, data = x)
    anova_res <- anova(art_model)
    anova_res_df <- as.data.frame(anova_res)
    return(anova_res_df)
  }
}

# Nest by target and trimester
aa <- full_data %>%
  group_by(target, trimester) %>%
  nest()

# Example to inspect first nested data frame
aa$data[[1]]

# Apply ART ANOVA to each nested dataframe
aa <- aa %>%
  mutate(ANOVA = map(data, doARTAnova))

# Inspect output
aa
aa$ANOVA[[1]]

anova_res_proteins <- aa %>%
  unnest("ANOVA") %>%
  select(-data) %>%
  clean_names()

# Adjust the p values
anova_res_proteins$p_adj_by <- p.adjust(anova_res_proteins$pr_f, method = "BY") # Benjamini–Yekutieli correction

anova_res_proteins %>%
  filter(p_adj_by < 0.05) %>%
  select(target) %>%
  distinct() # 2,187 significant targets vs 2,088 with BY correction

anova_res_proteins %>%
  filter(p_adj_by < 0.05, trimester == "Third") %>%
  select(target) %>%
  count(target) %>%
  distinct(target)

#write_csv(anova_res_proteins, "../output/tarca/statistics_ARTANOVA_results_proteins.csv")

# Next steps
posthoc <- aa %>%
  unnest(cols = c("target", "trimester", "data", "ANOVA")) %>%
  clean_names()
posthoc$p_adj_by <- p.adjust(posthoc$pr_f, method = "BY") # Benjamini–Yekutieli correction
#posthoc <- posthoc %>%
#  filter(p_adj_by < 0.05)
```

```{r}
# Run a post-hoc
doARTAnovaPostHoc <- function(x) {
  if(length(unique(x$group)) < 2) {
    return(NA)
  } else {
    art_model <- art(log2rfu ~ group, data = x)
    ph <- art.con(art_model, "group", adjust = "bonferroni")
    as.tibble(summary(ph))
  }
}

# Nest by target and trimester
aaph <- posthoc %>%
  group_by(target, trimester) %>%
  nest()

# Example to inspect first nested data frame
aaph$data[[1]]

# Apply the posthoc to each nested dataframe
aaph <- aaph %>%
  mutate(AnovaPH = map(data, doARTAnovaPostHoc))

# Inspect output
aaph
aaph$AnovaPH[[1]]

anovaph_res_proteins <- aaph %>%
  unnest(cols = c("target", "trimester", "AnovaPH")) %>%
  select(-data) %>%
  clean_names()

anovaph_res_proteins %>%
  filter(trimester == "Third", p_value < 0.05) %>%
  select(target) %>%
  count(target) %>%
  distinct(target)

#anovaph_res_proteins <- anovaph_res_proteins %>%
#  filter(p_value < 0.05) %>%
#  arrange(p_value)
?art.con
#write_csv(anovaph_res_proteins, "../output/tarca/statistics_ARTANOVA_posthoc_results_proteins.csv")
```
Output interpretation:

A positive *estimate* means the first group in the contrast (e.g., "preterm") has a higher aligned rank (hence, potentially higher transformed protein measurement) than the second ("term"). A negative value means the opposite.
-> shows the direction and size of the difference 

An example based on the first row: 
>On the aligned-rank scale, term pregnancies (controls) have, on average, an aligned rank 165.00 units lower than preterm pregnancies for Angiopoietin-1.

The *t.ratio* expresses how many standard errors the observed difference is away from zero. A larger absolute value means a more extreme difference (relative to noise), thus stronger evidence against the null hypothesis of no difference.
-> quantifies how strong the above difference is compared to its variability, forming the basis for statistical inference
>The observed group difference is (-)17.22 times the standard error — suggesting a robust effect.

-> The p value (4.037486e-47) indicates that this difference is statistically significant. 
```{r}
# Identify the 'double-validated' proteins i.e. significant in both KW and ANOVA
anovaph_res_proteins %>%
  filter(trimester == "Second", p_value < 0.05) %>%
  count(target) %>%
  unique() # 1024 in second trimester + 1040 in third trimester

dunn_res_proteins %>%
  filter(trimester == "Third", p.adj < 0.05) %>% 
  arrange(p.adj) %>%
  count(protein_name) %>%
  unique() # 1018 proteins in second trimester + 1036 proteins in third semester

# Compile target lists
anova_targets2 <- anovaph_res_proteins %>%
  filter(trimester == "Second") %>% 
  pull(target)
anova_targets3 <- anovaph_res_proteins %>%
  filter(trimester == "Third") %>% 
  pull(target)

dunn_target2 <- dunn_res_proteins %>%
  filter(trimester == "Second", p.adj < 0.05) %>%
  pull(protein_name)
dunn_target3 <- dunn_res_proteins %>%
  filter(trimester == "Third", p.adj < 0.05) %>%
  pull(protein_name)

# Calculate common targets for each trimester
common_targets2 <- intersect(anova_targets2, dunn_target2)
common_targets3 <- intersect(anova_targets3, dunn_target3)

# Add a conditional filter: match targets per trimester only to relevant intersection
anovaph_res_proteins_filtered <- anovaph_res_proteins %>%
  filter(
    (trimester == "Second" & target %in% common_targets2) |
    (trimester == "Third"  & target %in% common_targets3)
  )

dunn_res_proteins_filtered <- dunn_res_proteins %>%
  filter(
    (trimester == "Second" & protein_name %in% common_targets2) |
    (trimester == "Third"  & protein_name %in% common_targets3)
  )

anovaph_res_proteins_filtered
dunn_res_proteins_filtered

#write_csv(anovaph_res_proteins_filtered, "../output/tarca/statistics_filtered_DEGs_ANOVA.csv")
#write_csv(dunn_res_proteins_filtered, "../output/tarca/statistics_filtered_DEGs_Dunns.csv")
```

Check the proteomic data from the beginning again to see if all 
- if there is less than X counts in Y number of cells - then remove it

```{r}
library(ggplot2)
library(EnhancedVolcano)

# Sample data frame resembling results from ART ANOVA posthoc
set.seed(123)
results <- data.frame(
  feature = paste0("Feature", 1:100),
  estimate = rnorm(100, 0, 2),       # effect size (analogous to log2 fold change)
  t_ratio = rnorm(100, 0, 4),        # test statistic
  p_adj = runif(100, 0, 0.1)         # adjusted p-values
)

# Add -log10 adjusted p-value for ggplot
anova_volcano <- anovaph_res_proteins %>%
  filter(contrast == "Control - sPTD")
anova_volcano$neg_log10_padj <- -log10(anova_volcano$p_value)

### 1. Volcano plot using ggplot2 ###
ggplot(anova_volcano, aes(x = estimate, y = neg_log10_padj)) +
  geom_point(aes(color = p_value < 0.05 & abs(estimate) > 2), alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "grey", "TRUE" = "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  labs(
    title = "Volcano Plot (ggplot2) - Using Estimate and Adjusted p-value",
    x = "Effect Size (Estimate)",
    y = "-Log10 Adjusted p-value",
    color = "Significant"
  ) +
  theme_minimal()

### 2. Volcano plot using EnhancedVolcano ###
EnhancedVolcano(anova_volcano,
  lab = anova_volcano$target,
  x = "estimate",        # x-axis: effect size
  y = "p_value",           # y-axis: adjusted p-value
  pCutoff = 0.05,
  FCcutoff = 2,          # effect size cutoff
  title = "Volcano Plot (EnhancedVolcano)",
  subtitle = "ART ANOVA Posthoc Results",
  pointSize = 2.5,
  labSize = 4,
  col = c("grey30", "forestgreen", "royalblue", "red2"),
  legendPosition = "right"
)
```

## Generalised Linear Modelling
```{r}
full_data_wider <- full_data %>%
  pivot_wider(names_from = target, values_from = log2rfu) %>%
  select(-c(accession, geo_accession, ga_delivery))

second_trimester_wide <- second_trimester %>%
  pivot_wider(names_from = target, values_from = log2rfu) %>%
  select(-c(accession, geo_accession, ga_delivery))

third_trimester_wide <- third_trimester %>%
  pivot_wider(names_from = target, values_from = log2rfu) %>%
  select(-c(accession, geo_accession, ga_delivery))

# Check the baseline class
levels(full_data$group) # Control is baseline
```

*full_data_wider predictors*
$Control
[1] *"JAK2"* *"PCSK7"* *"Soggy-1"* "ULBP-2" "a2-Macroglobulin"
$PPROM
[1] *"IgE"*        
$sPTD
[1] "FGFR-3" "GPC5" "PTHrP" "Prothrombin"

*second trimester predictors*
$Control
[1] "H6ST1" "Hemoglobin" "IGFBP-1" "IGFBP-6" "Integrin a1b1" *"PCSK7"* *"Soggy-1"*      

$PPROM
[1] *"IgE"*        

$sPTD
[1] "HIPK3" *"JAK2"* "Livin B" "NOTC2" "PIM1" "Protein C" *"Prothrombin"*

*third trimester predictors*
$Control
[1] "Integrin a1b1" *"JAK2"* "Peroxiredoxin-6" *"Soggy-1"* "gpIIbIIIa"      

$PPROM
[1] *"IgE"*        

$sPTD
[1] "Apo A-I" "C1r" *"FGFR-3"* "IGFBP-3" "Livin B" *"PTHrP"* "TSH"

>increased levels of c-reactive protein and complement C5 in the preterm birth group ([Beernink et al., 2023](https://www.sciencedirect.com/science/article/pii/S0143400423001467)).

```{r}
# Full data
#x_proteins <- as.matrix(second_trimester_wide[, -c(1, 2)])
#x_trimesters <- model.matrix(~ trimester - 1, data = full_data_wider)
#x <- cbind(x_proteins, x_trimesters)
#y <- second_trimester_wide$group

# Second trimester
#x <- as.matrix(second_trimester_wide[, -c(1, 2)])
#y <- second_trimester_wide$group

# Third trimester
x <- as.matrix(third_trimester_wide[, -c(1, 2)])
y <- third_trimester_wide$group

fit <- glmnet(x, y, family = "multinomial")
plot(fit, xvar = "lambda", label = TRUE, type.coef = "2norm")

cv_fit <- cv.glmnet(x, y, family = "multinomial")
plot(cvfit)
best_lambda <- cv_fit$lambda.min
print(best_lambda)

final_model <- glmnet(x, y, family = "multinomial", lambda = best_lambda)
coef_list <- coef(final_model)

selected_vars <- lapply(coef_list, function(class_coef) {
  rownames(class_coef)[which(class_coef != 0)] # finds predictors with non-zero coefficients
})

print(selected_vars)
```
Non-zero coefficients: those selected by the LASSO penalty as relevant for distinguishing that class
- variables not listed (zero coefficient) are excluded by LASSO as non-informative (or redundant) for that class
- highlights potential biomarkers or features important for distinguishing that class

```{r}
coef_list[["Control"]] # log-odds ratios relative to the baseline class

get_selected_coefs <- function(coef_mat, class_name) {
  df <- as.data.frame(as.matrix(coef_mat))
  df$variable <- rownames(df)
  colnames(df)[1] <- "coefficient"
  
  # Filter non-zero coef excluding intercept
  df_sel <- df %>%
    filter(coefficient != 0 & variable != "(Intercept)") %>%
    mutate(class = class_name) %>%
    select(class, variable, coefficient)
  
  return(df_sel)
}

# Apply per class and combine
selected_coefs_df <- bind_rows(
  lapply(names(coef_list), function(cl) get_selected_coefs(coef_list[[cl]], cl))
)

selected_coefs_df <- selected_coefs_df %>%
  mutate(odds_ratio = exp(coefficient))

selected_coefs_df1 <- selected_coefs_df %>%
  dplyr::mutate(piodio = case_when(               # percentage increase or decrease in odds
    odds_ratio > 1 ~ (odds_ratio - 1) * 100,
    odds_ratio < 1 ~ (1 - odds_ratio) * 100
  )) %>%
  dplyr::mutate(ipp = (odds_ratio / 1 + odds_ratio) * 100) # implied percentage probability
print(selected_coefs_df1)
```
L1 regularization may help!

check the balance between l1 and l2 regularization

Odds-ratio interpretations:
OR ≈ 1: No change in odds with increase in predictor
OR > 1: Odds increase; predictor positively associated
OR < 1 (e.g., 0.5): Odds decrease; predictor negatively associated
OR ≈ 0 (e.g., 2e-63): Odds drastically decrease; near zero odds -> the predictor strongly reduces the odds of that class versus term

Concretely, if the coefficient for protein X under category sPTD is 0.5, it means that for every one-unit increase in protein X’s value, the log-odds of a sample being sPTD instead of term increase by 0.5.These log-odds can be converted to odds ratios by taking the exponential: exp(0.5) ≈ 1.65, meaning the odds of sPTD versus term are multiplied by about 1.65 for each unit increase in protein X.

-> they describe relative risk as log-odds changes, not absolute probabilities.

*PPROM	IgE	13.055350	4.675915e+05 (= 467,591.5)*
-> every one-unit increase in IgE log2RFUs, increases the odds of the pregnancy outcome being pPROM instead of term by a factor of 467,591.5
-> piodio: 46,759,050% increase in odds (???) that can't be right
-> ipp: 93,518,300% increase in implied probability? 

*sPTD	FGFR-3	-111.100550	5.618808e-49 (= 49 zeros in front of 5618808)* {no association with preterm birth}
-> every one-unit increase in FGFR-3 log2RFU, decreases the odds of the pregnancy outcome being sPTD (instead of term) by a factor of 5.618808e-49 (nearly zero and a near 100% decrease in odds)
-> piodio: 100% decrease in odds? yes, nearly
-> essentially this means that higher values of FGFR-3 in plasma dramatically reduce the odds of sPTD 

*sPTD	GPC5	-12.966021	2.338454e-06*
-> every one-unit increase in FGFR-3 log2RFU, decreases the odds of the pregnancy outcome being sPTD (instead of term) by a factor of 2.338454e-06 (nearly zero and a near 100% decrease in odds)
-> piodio: 99.99977% decrease in odds? these results are bold
-> as above but not as extreme, essentially

*sPTD PTHrP	-23.486765	6.307069e-11*
-> every one-unit increase in PTHrP log2RFU, decreases the odds of the pregnancy outcome being sPTD (instead of term) by a factor of 6.307069e-11 (nearly zero and a near 100% decrease in odds)
-> piodio: 100% decrease in odds again? bruh
-> let's see the pcaaaaaaa

*sPTD	Prothrombin	63.957288	5.974443e+27* {picked up in KEGG results in https://www.tandfonline.com/doi/full/10.1080/14767058.2023.2232074#d1e439}
-> every one-unit increase in Prothrombin log2RFUs, increases the odds of the pregnancy outcome being sPTD instead of term by a factor of 5.974443e+27
-> piodio: 5 × 10^29 % increase in odds? and we have a winner
 

# Statistical evaluation based on pathways
```{r}
kegg <- read_tsv("../data/tarca_et_al_2021/string/enrichment.KEGG.tsv")
go <- read_tsv("../data/tarca_et_al_2021/string/enrichment.GO.BioProcess.tsv")
```

```{r}
# Tidy the data
go <- go %>%
  clean_names() %>%
  select(-matching_proteins_in_your_network_i_ds) %>%
  filter(false_discovery_rate < 0.001)

go1 <- go %>%
  separate_rows(matching_proteins_in_your_network_labels, sep = ",") %>%
  rename(protein_name = matching_proteins_in_your_network_labels)
  
kegg <- kegg %>%
  clean_names() %>%
  select(-matching_proteins_in_your_network_i_ds) %>%
  filter(false_discovery_rate < 0.001)

kegg1 <- kegg %>%
  separate_rows(matching_proteins_in_your_network_labels, sep = ",") %>%
  rename(protein_name = matching_proteins_in_your_network_labels)
```

```{r}
# Join the datasets
go_joined <- right_join(go1, full_data, by = c("protein_name" = "target"))
go_joined %>%
  count(protein_name) # 1,165 proteins

kegg_joined <- right_join(kegg1, full_data, by = c("protein_name" = "target"))
kegg_joined %>%
  count(protein_name) # 1,165 proteins
```

```{r}
doKruskalTest <- function(x) {
  x %>%
    kruskal_test(log2rfu~group) %>%
  print(kruskalTest)
}

e <- go_joined %>%
  group_by(term_description, trimester) %>%
  nest()

e$data[[1]]

e <- go_joined %>%
  group_by(term_description, trimester) %>%
  nest() %>%
  mutate(Kruskal = map(data, doKruskalTest))

e
e$Kruskal[[1]]
e$data[[1]]

go_res_kruskal <- e %>%
  unnest(cols = c("term_description", "trimester", "Kruskal")) %>%
  select(-data)

# Adjust p values
go_res_kruskal$p_adj_bh <- p.adjust(go_res_kruskal$p, method = "BH") # Benjamini–Hochberg correction
go_res_kruskal$p_adj_by <- p.adjust(go_res_kruskal$p, method = "BY") # Benjamini–Yekutieli correction

go_res_kruskal %>%
  filter(p < 0.05) # 1963 significant overall 
go_res_kruskal %>%
  filter(p_adj_bh < 0.05) # 1907 significant with BH correction
go_res_kruskal %>%
  filter(p_adj_by < 0.05) # 1662 significant with BY correction

# Reorder the columns
go_res_kruskal <- go_res_kruskal %>%
  select(-method, everything(), method) %>%
  arrange(p_adj_by)

write_csv(go_res_kruskal, "../output/tarca/statistics_kruskal_wallis_results_GO.csv")

# Next steps
dunns3 <- e %>%
  unnest(cols = c("term_description", "trimester", "data", "Kruskal"))
dunns3$p_adj_by <- p.adjust(dunns3$p, method = "BY") # Benjamini–Yekutieli correction
```

```{r}
doDunnsTest <- function(x) {
    x %>%
    dunn_test(log2rfu~group) %>%
  print(dunnsTest)
  }

f <- dunns3 %>%
  group_by(term_description, trimester) %>%
  dplyr::filter(p_adj_by < 0.05) %>%
  nest()

f$data[[1]]

f <- dunns3 %>%
  group_by(term_description, trimester) %>%
  dplyr::filter(p_adj_by < 0.05) %>%
  nest() %>%
  mutate(Dunn = map(data, doDunnsTest))

f
f$Dunn[[1]]

go_res_dunn <- f %>%
  unnest(cols = c("term_description", "trimester", "Dunn")) %>%
  select(-data) %>%
  #drop_na(term_description) %>%
  arrange(p.adj)

write_csv(go_res_dunn, "../output/tarca/statistics_dunns_results_GO.csv")
```


# Hypergraph analysis
```{r}
hyp_data <- full_data %>%
  select(target, log2rfu, accession, trimester)

hyp_data <- hyp_data %>%
  pivot_wider(names_from = target, values_from = log2rfu)
```

```{r}
# Assign samples to rownames
hyp_data <- hyp_data %>%
  column_to_rownames("accession")

# Run hypergraph
runHypergraph <- function(sample, group, trimester) {
  
  sample_ids <- sample$accession
  samples_X <- hyp_data[rownames(hyp_data) %in% sample$accession, ]
  samples_X <- na.omit(samples_X)
  
  # Remove genes with SD = 0
  sd_scores <- apply(samples_X, 2, sd)
  samples_X_sd <- samples_X[, which(sd_scores > 0)]
  
  # Run correlation
  gene_cor <- cor(samples_X_sd, use = "pairwise.complete.obs", method = "spearman")
  
  # Binerise correlation values using SD of correlation matrix
  bin <- abs(gene_cor)
  bin[which(bin > sd(gene_cor))] <- 1
  bin[which(bin != 1)] <- 0 # hypergraph incidence matrix
  incidence_matrix = as.matrix(bin)

  # Generate hypergraph adjacency matrix by matrix multiplication
  hyp <- bin %*% t(bin) 
  adjacency_matrix = as.matrix(hyp)
  
  # Ranking
  #rowsum_res <- rowSums(hyp)
  #hist(rowsum_res) # non-parametric
  #row_sums_df <- as.data.frame(rowsum_res)
  #colnames(row_sums_df)[1] <- "rowsum"
  #row_sums_df$rank <- rank(row_sums_df$rowsum, ties.method = "average")
  #row_sums_df$norm_rank <- (row_sums_df$rank)/nrow(row_sums_df) # Divide by the number of rows
  
  # Add the outcome label
  #row_sums_df$group <- group
  #row_sums_df$trimester <- trimester
  
  # Sort ranks
  #row_sums_df <- row_sums_df[order(row_sums_df$rank, decreasing = TRUE), ]
  
  # Change rownames to column
  #row_sums_df$gene <- rownames(row_sums_df)

  #return(row_sums_df)
  return(adjacency_matrix)
}

e <- full_data %>%
  group_by(group, trimester) %>%
  nest() %>%
  mutate(hypergraph = pmap(list(data, group, trimester), runHypergraph))

e
#e$data[[1]]
e$hypergraph[[1]]

hyp_res <- e %>%
  mutate(hypergraph = map(hypergraph, ~ rownames_to_column(.x, var = "protein") %>% select(-c(group, trimester)))) %>%
  unnest(hypergraph) %>%
  select(-data)

hyp_res <- hyp_res %>%
  select(-gene)

# Import the full data for annotation
full_data2 <- read_csv("../data/tarca_et_al_2021/merged_normalised_datasets_for_modelling.csv")
full_data2 <- full_data2 %>%
  clean_names() %>%
  dplyr::select(target, group, entrez_gene_symbol, uni_prot) %>%
  distinct()

full_hyp_res <- right_join(hyp_res, full_data2, by = c("protein" = "target", "group" = "group"))
full_hyp_res2 <- full_hyp_res %>%
  dplyr::select(trimester, group, entrez_gene_symbol, uni_prot, protein, rowsum, rank, norm_rank) %>%
  unique()
```

```{r}
# Fix the missing data
# gene-names missing for: CS084, F229A, CG073, GNT2C, PKB a/b/g, PAL4D, ES1, CK094, UPK3L, PP13G, ANR63
library(biomaRt)
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
listAttributes(mart)[grep("uniprot", listAttributes(mart)$name), ]

# Connect to Ensembl via biomaRt
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Identify rows with NA gene symbols
na_indices <- which(is.na(full_hyp_res2$entrez_gene_symbol))
uniprot_ids_to_lookup <- unique(full_hyp_res2$uni_prot[na_indices])

# Query Ensembl: Map UniProt ID to gene symbol
gene_mapping <- getBM(
  attributes = c("uniprotswissprot", "external_gene_name"),
  filters = "uniprotswissprot",
  values = uniprot_ids_to_lookup,
  mart = mart
)

# Fill in the missing gene symbols
for (i in na_indices) {
  matching_gene <- gene_mapping$external_gene_name[gene_mapping$uniprotswissprot == full_hyp_res2$uni_prot[i]]
  if (length(matching_gene) > 0) {
    full_hyp_res2$entrez_gene_symbol[i] <- matching_gene[1]
  }
}

# View the updated dataframe
full_hyp_res2 %>%
  filter(protein == "PKB a/b/g") # GNT2C, PKB a/b/g, and ES1 still NA

# Manually add the ones still missing
# Manually fix missing gene symbols
full_hyp_res2$entrez_gene_symbol[full_hyp_res2$uni_prot == "Q8NFS9"] <- "GCNT2"
full_hyp_res2$entrez_gene_symbol[full_hyp_res2$protein == "PKB a/b/g"] <- "AKT1 AKT2 AKT3"
full_hyp_res2$entrez_gene_symbol[full_hyp_res2$uni_prot == "P30042"] <- "GATD3"

#write_csv(full_hyp_res2, "../output/tarca/full_hypergraph_results.csv")
```


```{r}
# Heatmap visualisations
adj_matrix_ctrl <- e$hypergraph[[1]]
euc_dist <- dist(adj_matrix_ctrl)
hc <- hclust(euc_dist, method = "complete")

png("../output/control_heatmap.png", width = 1650, height = 1650, res = 300)
heatmap(adj_matrix_ctrl, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc),
        symm = TRUE, scale = "none", col = heat.colors(256))
dev.off()

adj_matrix_sptd <- e$hypergraph[[2]]
euc_dist <- dist(adj_matrix_ctrl)
hc <- hclust(euc_dist, method = "complete")

png("../output/sPTD_heatmap.png", width = 1600, height = 1600, res = 300)
heatmap(adj_matrix_sptd, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc),
        symm = TRUE, scale = "none", col = heat.colors(256))
dev.off()

adj_matrix_pprom <- e$hypergraph[[3]]
euc_dist <- dist(adj_matrix_ctrl)
hc <- hclust(euc_dist, method = "complete")

png("../output/pprom_heatmap.png", width = 1600, height = 1600, res = 300)
heatmap(adj_matrix_pprom, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc),
        symm = TRUE, scale = "none", col = heat.colors(256))
dev.off()
```

```{r}
library(pheatmap)
pheatmap(adj_matrix_ctrl,
         clustering_distance_rows = euc_dist,
         clustering_distance_cols = euc_dist,
         clustering_method = "complete",
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = colorRampPalette(c("white", "blue"))(256),
         filename = "../output/control_pheatmap.png")

pheatmap(adj_matrix_sptd,
         clustering_distance_rows = euc_dist,
         clustering_distance_cols = euc_dist,
         clustering_method = "complete",
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = colorRampPalette(c("white", "blue"))(256),
         filename = "../output/sPTD_pheatmap.png")

pheatmap(adj_matrix_pprom,
         clustering_distance_rows = euc_dist,
         clustering_distance_cols = euc_dist,
         clustering_method = "complete",
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = colorRampPalette(c("white", "blue"))(256),
         filename = "../output/pPROM_pheatmap.png")
```

The proteome is completely screwed in pprom
- It changes a little bit in sPTD but in PPROM it gets all over the place!
- completely dysfunctional
-> however, is that because of the pprom per se, or due to an underlying infection (i.e. bacteremia)?
-> the exclusion criteria is not presented in the original publication so it's unclear if the women with pprom had any underlying conditions

```{r}
adj_matrix_ctrl2 <- e$hypergraph[[1]]
euc_dist1 <- dist(adj_matrix_ctrl2)
hc1 <- hclust(euc_dist1, method = "complete")

adj_matrix_sptd2 <- e$hypergraph[[2]]
euc_dist2 <- dist(adj_matrix_sptd2)
hc2 <- hclust(euc_dist2, method = "complete")

adj_matrix_pprom2 <- e$hypergraph[[3]]
euc_dist3 <- dist(adj_matrix_pprom2)
hc3 <- hclust(euc_dist3, method = "complete")

adj_matrix_ctrl3 <- e$hypergraph[[4]]
euc_dist4 <- dist(adj_matrix_ctrl3)
hc4 <- hclust(euc_dist4, method = "complete")

adj_matrix_sptd3 <- e$hypergraph[[5]]
euc_dist5 <- dist(adj_matrix_sptd3)
hc5 <- hclust(euc_dist5, method = "complete")

adj_matrix_pprom3 <- e$hypergraph[[6]]
euc_dist6 <- dist(adj_matrix_pprom3)
hc6 <- hclust(euc_dist6, method = "complete")

library(pheatmap)
phc2 <- pheatmap(adj_matrix_ctrl2,
         clustering_distance_rows = euc_dist1,
         clustering_distance_cols = euc_dist1,
         clustering_method = "complete",
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = colorRampPalette(c("white", "blue"))(256),
         filename = "../output/pheatmap_ctrl2.png")
phc3 <- pheatmap(adj_matrix_ctrl3,
         clustering_distance_rows = euc_dist4,
         clustering_distance_cols = euc_dist4,
         clustering_method = "complete",
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = colorRampPalette(c("white", "blue"))(256),
         filename = "../output/pheatmap_ctrl3.png")
phsptd2 <- pheatmap(adj_matrix_sptd2,
         clustering_distance_rows = euc_dist2,
         clustering_distance_cols = euc_dist2,
         clustering_method = "complete",
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = colorRampPalette(c("white", "blue"))(256),
         filename = "../output/pheatmap_sPTD2.png")
phsptd3 <- pheatmap(adj_matrix_sptd3,
         clustering_distance_rows = euc_dist5,
         clustering_distance_cols = euc_dist5,
         clustering_method = "complete",
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = colorRampPalette(c("white", "blue"))(256),
         filename = "../output/pheatmap_sPTD3.png")
pheatmap(adj_matrix_pprom2,
         clustering_distance_rows = euc_dist3,
         clustering_distance_cols = euc_dist3,
         clustering_method = "complete",
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = colorRampPalette(c("white", "blue"))(256),
         filename = "../output/pheatmap_pPROM2.png")
pheatmap(adj_matrix_pprom3,
         clustering_distance_rows = euc_dist6,
         clustering_distance_cols = euc_dist6,
         clustering_method = "complete",
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = colorRampPalette(c("white", "blue"))(256),
         filename = "../output/pheatmap_pPROM3.png")
```

*Central clusters*
```{r}
#phc2, phc3, phsptd2, phsptd3
# Control: Second trimester
central_cluster_ctrl2 <- cutree(phc2$tree_row, k = 2)
central_cluster_ctrl2 <- as.data.frame(central_cluster_ctrl2) %>%
  rownames_to_column(var = "target") 
# Which is the smaller cluster? 1
central_cluster_ctrl2 <- central_cluster_ctrl2 %>%
  #count(central_cluster_ctrl2)
  filter(central_cluster_ctrl2 == 1)

# Control: Third trimester
central_cluster_ctrl3 <- cutree(phc3$tree_row, k = 2)
central_cluster_ctrl3 <- as.data.frame(central_cluster_ctrl3) %>%
  rownames_to_column(var = "target")
# Which is the smaller cluster? 1
central_cluster_ctrl3 <- central_cluster_ctrl3 %>%
  #count(central_cluster_ctrl3)
  filter(central_cluster_ctrl3 == 1)

# sPTD: Second trimester
central_cluster_sptd2 <- cutree(phsptd2$tree_row, k = 4)
central_cluster_sptd2 <- as.data.frame(central_cluster_sptd2) %>%
  rownames_to_column(var = "target") 
# Which is the smallest cluster? 1
central_cluster_sptd2 <- central_cluster_sptd2 %>%
  #count(central_cluster_sptd2) %>%
  filter(central_cluster_sptd2 == 1)

# sPTD: Third trimester
central_cluster_sptd3 <- cutree(phsptd3$tree_row, k = 2)
central_cluster_sptd3 <- as.data.frame(central_cluster_sptd3) %>%
  rownames_to_column(var = "target") 
# Which is the smaller cluster? 1
central_cluster_sptd3 <- central_cluster_sptd3 %>%
  #count(central_cluster_sptd3) %>%
  filter(central_cluster_sptd3 == 1)
```

```{r}
# Which proteins are overlapping in the second trimester?
second_trimester_overlap <- intersect(central_cluster_ctrl2$target, central_cluster_sptd2$target)
second_trimester_overlap

# Create a Venn diagram
n_cc_sptd2 <- length(central_cluster_sptd2$target)
n_cc_ctrl2 <- length(central_cluster_ctrl2$target)
n_overlap2 <- length(second_trimester_overlap)

venn_data <- Venn(SetNames = c("Control", "sPTD"),
                  Weight = c(
                    "10" = n_cc_ctrl2,
                    "01" = n_cc_sptd2,
                    "11" = n_overlap2
                  ))
# Modify colors for sets and overlaps:
venn_drawing <- compute.Venn(venn_data, doWeights = FALSE)
gp <- VennThemes(venn_drawing)
gp[["Set"]][["Control"]]$fill <- "#E64B3599"   
gp[["Set"]][["sPTD"]]$fill <- "#4DBBD599"      
gp[["Face"]][["11"]]$fill <- "#8491B499"       
gp[["Set"]][["Control"]]$col <- "#E64B35FF"
gp[["Set"]][["sPTD"]]$col <- "#4DBBD5FF"
grid.newpage()
png("venn_second_trimester.png", width = 600, height = 600)
plot(venn_drawing, gpList = gp)
dev.off()

# Which proteins are sPTD specific?
cc_sptd2_specific <- central_cluster_sptd2 %>%
  filter(!target %in% second_trimester_overlap)
cc_sptd2_specific

# Which proteins are term specific?
cc_ctrl2_specific <- central_cluster_ctrl2 %>%
  filter(!target %in% second_trimester_overlap)
cc_ctrl2_specific
```

```{r}
cc_sptd2_specific$trimester = "Second"
cc_sptd2_specific$central_cluster = "sPTD-specific"

cc_ctrl2_specific$trimester = "Second"
cc_ctrl2_specific$central_cluster = "term-specific"

central_clusters2 <- full_join(cc_sptd2_specific, cc_ctrl2_specific) %>%
  select(-c(central_cluster_sptd2, central_cluster_ctrl2)) # number of rows should be 404
```


```{r}
# Which proteins are overlapping in the second trimester?
third_trimester_overlap <- intersect(central_cluster_ctrl3$target, central_cluster_sptd3$target)
third_trimester_overlap

# Create a Venn diagram
n_cc_sptd3 <- length(central_cluster_sptd3$target)
n_cc_ctrl3 <- length(central_cluster_ctrl3$target)
n_overlap3 <- length(third_trimester_overlap)

# Which proteins are sPTD specific?
cc_sptd3_specific <- central_cluster_sptd3 %>%
  filter(!target %in% third_trimester_overlap)
cc_sptd3_specific

# Which proteins are term specific?
cc_ctrl3_specific <- central_cluster_ctrl3 %>%
  filter(!target %in% third_trimester_overlap)
cc_ctrl3_specific

516 - 255
```
```{r}
cc_sptd3_specific$trimester = "Third"
cc_sptd3_specific$central_cluster = "sPTD-specific"

cc_ctrl3_specific$trimester = "Third"
cc_ctrl3_specific$central_cluster = "term-specific"

central_clusters3 <- full_join(cc_sptd3_specific, cc_ctrl3_specific) %>%
  select(-c(central_cluster_ctrl3, central_cluster_sptd3))
216 + 261 # number of rows should be 477

central_clusters <- full_join(central_clusters2, central_clusters3) # should have 881 rows
write_csv(central_clusters, "../output/tarca/central_clusters.csv")
```


# Biomarker interactions
```{r}
string_interactions <- read_tsv("../data/tarca_et_al_2021/string/string_interactions.tsv")
string_interactions
```

```{r}
potential_biomarkers <- c("AGER", "APOB", "PTHLH", "YES1", "LAG3", "A2M", "LY86", "CDON", "SERPINA5", "CSNK2A1", "APP", "PLAUR", "PDE11A", "PCSK7", "EIF5", "SORCS2", "ITIH4", "IMPDH2", "JAK2")
markers <- full_hyp_res %>%
  filter(entrez_gene_symbol %in% potential_biomarkers)

string_interactions %>%
  clean_names() %>%
  filter(number_node1 %in% potential_biomarkers & node2 %in% potential_biomarkers) %>%
  select(-c(node1_string_id, node2_string_id)) %>%
  arrange(desc(combined_score))
```


